---
title: "Dynamic Surface Water Extent (DSWE) Data for the Klamath Basin"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dynamic Surface Water Extent (DSWE) Data for the Klamath Basin}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

This vignette documents the **Dynamic Surface Water Extent Science (DSWE)** dataset
for the **Klamath Basin**, derived from the **Landsat Collection 2 Level-3 DSWE
Science Product** using the official USGS DSWE algorithm.

The dataset summarizes surface water presence across seven wildlife refuges in the
Klamath Basin using daily Landsat imagery processed through **Google Earth Engine
(GEE)** and accessed via the R package **rgee**.

The analysis focuses on **water presence classes 1, 2, and 3**, which together
characterize inundation extent across a range of detection confidence levels.

# DSWE Product Description

The DSWE product provides three interpreted outputs, each with increasing levels of
filtering:

- **INTR** – Interpreted (raw)
- **INTSM** – Interpreted with spectral masks
- **INWAM** – Interpreted with all masks applied

This dataset uses the **INWAM** layer, the most refined DSWE product, which applies
comprehensive filtering for clouds, cloud shadows, terrain effects, and non-water
land cover.

# INWAM Pixel Value Interpretation

The INWAM layer classifies each pixel using values from 0–255. Water presence is
indicated by pixel values **1, 2, and 3**, which are the focus of this dataset.

## Table 1. INWAM Layer Categories

| Pixel Value | Interpretation                      | Water Confidence / Classification |
|------------|--------------------------------------|----------------------------------|
| 0          | Not water                            | No water detected or masked      |
| 1          | Water – high confidence              | 4–5 diagnostic tests true        |
| 2          | Water – moderate confidence          | 3 diagnostic tests true          |
| 3          | Partial surface water (conservative) | Conservative wetland detection   |
| 4          | Partial surface water (aggressive)   | Low confidence detection         |
| 9          | Cloud / cloud shadow                 | QA-masked                        |
| 255        | Fill                                 | No data                           |

Only pixel values **1, 2, and 3** are retained in the summarized dataset.

# Study Area: Klamath Basin Water Refuges

Surface water was summarized for the following seven areas, defined using bounding
boxes.

## Table 2. Refuges and Bounding Boxes

| Refuge Name                 | Bounding Box (lat, lon) |
|----------------------------|--------------------------|
| Upper Klamath Lake          | 42.225167, −122.138907 → 42.614644, −121.770893 |
| Clear Lake                  | 41.781010, −121.289207 → 41.995292, −121.015256 |
| Tule Lake                   | 41.800411, −121.598288 → 41.995232, −121.347044 |
| Bear Valley                 | 42.034977, −121.979966 → 42.096687, −121.889353 |
| Lower Klamath + Sheepy Lake | 41.883384, −121.851156 → 42.033576, −121.591674 |
| Klamath Marsh               | 42.827148, −121.806202 → 43.072530, −121.517889 |
| Broad Area (entire basin)   | 41.737336, −122.136971 → 42.677414, −121.012918 |

Bounding boxes were defined using `bboxfinder.com`.

# Data Access Requirements

## Google Earth Engine

Accessing this dataset **requires a Google account** with permission to use
**Google Earth Engine**.

## Required R Packages

```{r eval=FALSE}
library(rgee)
library(dplyr)
library(lubridate)
```

Before running any Earth Engine operations, users must authenticate:

```{r eval=FALSE}
rgee::ee_Initialize()
```

# Data Processing Workflow

1. Landsat Collection 2 imagery was accessed via Google Earth Engine.
2. Daily DSWE classifications were computed using the USGS DSWE algorithm.
3. INWAM pixel values **1, 2, and 3** were extracted.
4. Cloud-contaminated scenes were excluded.
5. Daily results were aggregated to monthly summaries.
6. Percent surface water coverage was calculated per refuge.

## Temporal Coverage

- **Start date:** 2016-10-01  
- **End date:** 2023-08-31  

Not all months contain observations due to cloud cover and Landsat overpass
availability.

# Output Structure

The final dataset contains monthly summaries with the following variables:

| Column Name        | Description |
|--------------------|-------------|
| area               | Refuge or analysis area |
| year               | Calendar year |
| month              | Calendar month |
| mean_total_water   | Mean % area with Class 1–3 water |
| sd_total_water     | Standard deviation of total water |
| mean_class1        | Mean % high-confidence water |
| mean_class2        | Mean % moderate-confidence water |
| mean_class3        | Mean % partial surface water |
| n                  | Number of cloud-free Landsat images |

All percentage values range from **0–100**.

# Results Summary (2017–2020)

Monthly summaries show strong seasonal variability across refuges, with:

- Near-complete inundation during winter months in some years
- Persistent low-level water presence in summer months
- High interannual variability driven by hydrology and management practices

These summaries support comparative analyses of wetland extent, seasonal inundation
patterns, and long-term trends across the Klamath Basin.

# Limitations and Notes

- Only Landsat scenes with minimal cloud cover are included.
- DSWE Class 4 (aggressive partial water) is excluded to reduce false positives.
- Bulk USGS DSWE downloads and non-R implementations were not evaluated.
- Results reflect surface water presence, not depth or volume.

# Citation

Users should cite the USGS DSWE algorithm, Landsat Collection 2, and Google Earth
Engine when using this dataset.
