---
title: "Flow Data"
author: "Badhia Yunes Katz"
date: "August 2026"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.width = 8, fig.height = 5)

library(klamathWaterData)
library(dplyr)
library(ggplot2)
library(leaflet)
library(knitr)
library(RColorBrewer)

colors_full <-  c("#9A8822", "#F5CDB4", "#F8AFA8", "#FDDDA0", "#74A089", #Royal 2
                  "#899DA4", "#C93312", "#DC863B", # royal 1 (- 3)
                  "#F1BB7B", "#FD6467", "#5B1A18", # Grand Budapest 1 (-4)
                  "#D8B70A", "#02401B", "#A2A475", # Cavalcanti 1
                  "#E6A0C4", "#C6CDF7", "#D8A499", "#7294D4", #Grand Budapest 2
                  "#9986A5", "#EAD3BF", "#AA9486", "#B6854D", "#798E87" # Isle of dogs 2 altered slightly
)
```

## Overview

This vignette summarizes `flow_data`, the packaged flow dataset for the Klamath Basin
watershed, along with its companion metadata table `flow_gage`. For each stream
grouping, a date-coverage plot and a summary of the available date range are provided
to give an easy visual reference for data availability and to highlight gaps.


`flow_data` columns:

  - `location`: stream/waterbody associated with the gage
  - `gage_name`: name of the gaging station
  - `gage_id`: unique identifier for the gage
  - `variable_name`: variable measured ("flow")
  - `value`: observed flow value (numeric)
  - `unit`: units of measurement (cfs)
  - `statistic`: summary statistic type ("mean")
  - `date`: date of the observation

`flow_gage` columns:

  - `location`, `gage_name`, `gage_id`: same as above
  - `agency`: agency responsible for operating or maintaining the gage
  - `latitude`, `longitude`: gage location
  - `river_mile`: closest river mile relative to river mouth or a reference point
  - `huc8`: 8-digit Hydrologic Unit Code identifying the sub-basin where the gage is located

```{r, include=FALSE}
# an additional grouping for the upper basin tributaries and the handful of
# gages in flow_gage that don't currently have a `location` assigned
stream_groups <- list(
  "Klamath River (mainstem)" = c("klamath river"),
  "Trinity River" = c("trinity river", "sf trinity river"),
  "Scott River" = c("scott river", "shackleford creek", "mill creek", "middle creek",
                     "kelsey creek", "canyon creek", "boulder creek", "tompkins creek",
                     "willow creek"),
  "Shasta River" = c("shasta river"),
  "Salmon River" = c("salmon river"),
  "Indian Creek" = c("indian creek"),
  "Link, Williamson & Sprague Basin" = c("link river", "williamson river", "sprague river",
                                          "north fork sprague river", "south fork sprague river",
                                          "sycan river")
)

# gages in flow_gage with no `location` assigned (e.g. Oregon canal/drain sites) -
# identified by gage_id since `location` is NA for these rows
unassigned_gage_ids <- flow_gage |>
  filter(is.na(location)) |>
  pull(gage_id)

# a small, high-contrast qualitative palette (one hue per stream group) - the
# muted colors_full palette above works fine for coloring individual gages
# within a facet, but its similar-looking tones make it hard to tell stream
# groups apart at a glance on the map, so groups get their own palette instead
group_names <- c(names(stream_groups), "Other / unassigned")
group_colors <- setNames(RColorBrewer::brewer.pal(length(group_names), "Dark2"), group_names)
```

## All flow gages

```{r}
group_lookup <- stream_groups |>
  purrr::imap(\(locations, group) tibble(location = locations, group = group)) |>
  bind_rows()

gage_map_data <- flow_gage |>
  left_join(group_lookup, by = "location") |>
  mutate(group = ifelse(gage_id %in% unassigned_gage_ids, "Other / unassigned", group)) |>
  filter(!is.na(latitude), !is.na(longitude))

group_pal <- colorFactor(palette = group_colors[group_names], domain = gage_map_data$group)

leaflet(gage_map_data) |>
  addTiles() |>
  addCircleMarkers(
    lng = ~longitude, lat = ~latitude,
    color = ~group_pal(group),
    radius = 5, stroke = FALSE, fillOpacity = 0.8,
    label = ~paste0(gage_name, " (", gage_id, ")")
  ) |>
  addLegend(position = "bottomright", pal = group_pal, values = ~group, title = "Stream group")
```

## Flow by stream {.tabset .tabset-fade .tabset-pills}

```{r, include=FALSE}
# shared helper used by every tab below: filter flow_data/flow_gage down to one
# stream group, plot date coverage per gage, and summarize the date range per gage
flow_group_data <- function(locations) {
  flow_data |> filter(location %in% locations)
}

flow_group_gages <- function(locations) {
  flow_gage |> filter(location %in% locations)
}

plot_date_coverage <- function(df) {
  df |> ggplot(aes(x = date, y = value, color = gage_name)) +
    geom_line() +
    scale_color_manual(values = colors_full) +
    facet_wrap(~gage_name) +
    theme_minimal() +
    theme(legend.position = "none", axis.text.x = element_text(angle = 90, hjust = 1)) +
    labs(title = "Flow vs Date by Gage", x = "Date", y = "Flow (cfs)")
}

summarize_date_range <- function(df) {
  df |>
    group_by(gage_name, gage_id) |>
    summarize(
      min_date = min(date),
      max_date = max(date),
      record_count = n(),
      .groups = "drop"
    ) |>
    arrange(gage_name)
}
```

### Klamath River (mainstem)

```{r}
kable(flow_group_gages(stream_groups[["Klamath River (mainstem)"]]) |>
        select(gage_name, gage_id, agency, river_mile))
```

#### Date coverage

```{r}
plot_date_coverage(flow_group_data(stream_groups[["Klamath River (mainstem)"]]))
```

#### Date range summary

```{r}
kable(summarize_date_range(flow_group_data(stream_groups[["Klamath River (mainstem)"]])))
```

### Trinity River

```{r}
kable(flow_group_gages(stream_groups[["Trinity River"]]) |>
        select(gage_name, gage_id, agency, river_mile))
```

#### Date coverage

```{r}
plot_date_coverage(flow_group_data(stream_groups[["Trinity River"]]))
```

#### Date range summary

```{r}
kable(summarize_date_range(flow_group_data(stream_groups[["Trinity River"]])))
```

### Scott River

Includes the mainstem Scott River gages along with the smaller Quartz Valley
Indian Community-monitored tributaries (Shackleford, Mill, Middle, Kelsey, Canyon,
Boulder, and Tompkins Creeks, and Willow Creek).

```{r}
kable(flow_group_gages(stream_groups[["Scott River"]]) |>
        select(gage_name, gage_id, agency, river_mile))
```

#### Date coverage

```{r}
plot_date_coverage(flow_group_data(stream_groups[["Scott River"]]))
```

#### Date range summary

```{r}
kable(summarize_date_range(flow_group_data(stream_groups[["Scott River"]])))
```

### Shasta River

```{r}
kable(flow_group_gages(stream_groups[["Shasta River"]]) |>
        select(gage_name, gage_id, agency, river_mile))
```

#### Date coverage

```{r}
plot_date_coverage(flow_group_data(stream_groups[["Shasta River"]]))
```

#### Date range summary

```{r}
kable(summarize_date_range(flow_group_data(stream_groups[["Shasta River"]])))
```

### Salmon River

```{r}
kable(flow_group_gages(stream_groups[["Salmon River"]]) |>
        select(gage_name, gage_id, agency, river_mile))
```

#### Date coverage

```{r}
plot_date_coverage(flow_group_data(stream_groups[["Salmon River"]]))
```

#### Date range summary

```{r}
kable(summarize_date_range(flow_group_data(stream_groups[["Salmon River"]])))
```

### Indian Creek

```{r}
kable(flow_group_gages(stream_groups[["Indian Creek"]]) |>
        select(gage_name, gage_id, agency, river_mile))
```

#### Date coverage

```{r}
plot_date_coverage(flow_group_data(stream_groups[["Indian Creek"]]))
```

#### Date range summary

```{r}
kable(summarize_date_range(flow_group_data(stream_groups[["Indian Creek"]])))
```

### Link, Williamson & Sprague Basin

Upper basin tributaries: Link River, Williamson River, and the Sprague River
system (mainstem, North Fork, South Fork, and Sycan River).

```{r}
kable(flow_group_gages(stream_groups[["Link, Williamson & Sprague Basin"]]) |>
        select(gage_name, gage_id, agency, river_mile))
```

#### Date coverage

```{r}
plot_date_coverage(flow_group_data(stream_groups[["Link, Williamson & Sprague Basin"]]))
```

#### Date range summary

```{r}
kable(summarize_date_range(flow_group_data(stream_groups[["Link, Williamson & Sprague Basin"]])))
```

### Other / unassigned gages

`flow_gage` includes a some of stations that are not located on a stream, therefore `location`
is unassigned (mostly Oregon canal/drain sites near Klamath Falls, plus one Scott Valley
reservation site). They're listed here by `gage_id` rather than by stream grouping.

```{r}
kable(flow_gage |> filter(gage_id %in% unassigned_gage_ids) |>
        select(gage_name, gage_id, agency, river_mile))
```

#### Date coverage

```{r}
flow_data |>
  filter(gage_id %in% unassigned_gage_ids) |>
  ggplot(aes(x = date, y = value, color = gage_name)) +
  geom_line() +
  scale_color_manual(values = colors_full) +
  facet_wrap(~gage_name) +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Flow vs Date by Gage", x = "Date", y = "Flow (cfs)")
```

#### Date range summary

```{r}
flow_data |>
  filter(gage_id %in% unassigned_gage_ids) |>
  group_by(gage_name, gage_id) |>
  summarize(
    min_date = min(date),
    max_date = max(date),
    record_count = n(),
    .groups = "drop"
  ) |>
  arrange(gage_name) |>
  kable()
```
