---
title: "Flow Data"
author: "Badhia Yunes Katz"
date: "August 2026"
output: html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.width = 8, fig.height = 5)

library(klamathWaterData)
library(dplyr)
library(ggplot2)
library(leaflet)
library(knitr)
library(RColorBrewer)

colors_full <-  c("#9A8822", "#F5CDB4", "#F8AFA8", "#FDDDA0", "#74A089", #Royal 2
                  "#899DA4", "#C93312", "#DC863B", # royal 1 (- 3)
                  "#F1BB7B", "#FD6467", "#5B1A18", # Grand Budapest 1 (-4)
                  "#D8B70A", "#02401B", "#A2A475", # Cavalcanti 1
                  "#E6A0C4", "#C6CDF7", "#D8A499", "#7294D4", #Grand Budapest 2
                  "#9986A5", "#EAD3BF", "#AA9486", "#B6854D", "#798E87" # Isle of dogs 2 altered slightly
)
```

## Overview

This vignette summarizes `flow_data`, the packaged flow dataset for the Klamath Basin
watershed, along with its companion metadata table `flow_gage`. Gages are grouped by
their 8-digit Hydrologic Unit Code (`huc8`) sub-basin - each tab below is one HUC8
sub-basin. For each sub-basin, a date-coverage plot and a summary of the available date range are provided
to give an easy visual reference for data availability and to highlight gaps.

```{r, include = FALSE}
classify_frequency <- function(median_gap_days) {
  case_when(
    median_gap_days <= 2 ~ "daily",
    median_gap_days <= 10 ~ "weekly",
    median_gap_days <= 20 ~ "biweekly (roughly twice a month)",
    median_gap_days <= 40 ~ "monthly",
    TRUE ~ "irregular"
  )
}

# gage_ids present in flow_data but missing from flow_gage get an explicit
# label here instead of silently becoming NA - see the callout below the
# source table
source_summary <- flow_data |>
  left_join(flow_gage |> select(gage_id, agency), by = "gage_id") |>
  mutate(agency = ifelse(is.na(agency), "USBR (gage not yet in flow_gage metadata)", agency)) |>
  group_by(agency) |>
  summarize(
    n_gages = n_distinct(gage_id),
    n_records = n(),
    min_date = min(date),
    max_date = max(date),
    median_gap_days = median(as.numeric(unlist(sapply(split(date, gage_id), \(x) diff(sort(unique(x)))))), na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(typical_frequency = classify_frequency(median_gap_days)) |>
  select(-median_gap_days) |>
  arrange(desc(n_records))

unmatched_gage_ids <- setdiff(unique(flow_data$gage_id), flow_gage$gage_id)
```

`flow_data` covers **`r format(min(flow_data$date), "%B %Y")` through
`r format(max(flow_data$date), "%B %Y")`**, with `r format(nrow(flow_data), big.mark = ",")`
records from `r n_distinct(flow_data$gage_id)` gages.

Data sources include: 

- USGS (via NWIS daily values)
- Water Quality Exchange / Water Quality Portal (WQX)
- U.S. Bureau of Reclamation


Data collection frequency:

- USGS gages report daily mean flow,
- WQX-sourced tribal sites are periodic grab samples roughly twice a month:

```{r}
source_summary <- flow_data |>
  left_join(flow_gage |> select(gage_id, agency), by = "gage_id") |>
  mutate(
    agency = ifelse(
      is.na(agency),
      "USBR (gage not yet in flow_gage metadata)",
      agency
    )
  ) |>
  group_by(agency) |>
  summarize(
    gage_ids = paste(sort(unique(gage_id)), collapse = ", "),
    n_gages = n_distinct(gage_id),
    n_records = n(),
    min_date = min(date),
    max_date = max(date),
    median_gap_days = median(
      as.numeric(
        unlist(
          sapply(
            split(date, gage_id),
            \(x) diff(sort(unique(x)))
          )
        )
      ),
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  mutate(
    typical_frequency = classify_frequency(median_gap_days)
  ) |>
  select(-median_gap_days) |>
  arrange(desc(n_records))

kable(
  source_summary,
  col.names = c(
    "Source / agency",
    "Gage IDs",
    "Gages",
    "Records",
    "Earliest date",
    "Latest date",
    "Typical frequency"
  )
)
```

Note: The note below will get updated when joined to main since flow_data changes have recently been
merged into main form a different branch. This comment should be resolved once this branch is joined.
Keeping as a placeholder. 

`r if (length(unmatched_gage_ids) > 0) paste0("**Note:** ", length(unmatched_gage_ids), " gage_id(s) present in flow_data (", paste(unmatched_gage_ids, collapse = ", "), ") have no matching row in flow_gage, so they carry no agency/location/huc8/coordinates and do not appear on the map or in any sub-basin tab below.")`

`flow_data` columns:

  - `location`: stream/waterbody associated with the gage
  - `gage_name`: name of the gaging station
  - `gage_id`: unique identifier for the gage
  - `variable_name`: variable measured ("flow")
  - `value`: observed flow value (numeric)
  - `unit`: units of measurement (cfs)
  - `statistic`: summary statistic type ("mean")
  - `date`: date of the observation

`flow_gage` columns:

  - `location`, `gage_name`, `gage_id`: same as above
  - `agency`: agency responsible for operating or maintaining the gage
  - `latitude`, `longitude`: gage location
  - `river_mile`: closest river mile relative to river mouth or a reference point
  - `huc8`: 8-digit Hydrologic Unit Code identifying the sub-basin where the gage is located

```{r, include=FALSE}
# HUC8 -> official sub-basin name, for every HUC8 that appears in flow_gage.
huc8_basin_names <- c(
  "18010201" = "Williamson",
  "18010202" = "Sprague",
  "18010203" = "Upper Klamath Lake",
  "18010204" = "Lost",
  "18010205" = "Butte",
  "18010206" = "Upper Klamath",
  "18010207" = "Shasta",
  "18010208" = "Scott",
  "18010209" = "Lower Klamath",
  "18010210" = "Salmon",
  "18010211" = "Trinity",
  "18010212" = "South Fork Trinity"
)

basin_label <- function(huc8) {
  huc8_chr <- as.character(huc8)
  name <- huc8_basin_names[huc8_chr]
  ifelse(is.na(name), paste("HUC8", huc8_chr), paste0(name, " (HUC8 ", huc8_chr, ")"))
}

huc8_codes <- flow_gage |>
  filter(!is.na(huc8)) |>
  distinct(huc8) |>
  arrange(huc8) |>
  pull(huc8)

# one group per sub-basin actually present in flow_gage, named list of huc8 codes
stream_groups <- setNames(as.list(huc8_codes), basin_label(huc8_codes))

# a high-contrast qualitative palette (one hue per sub-basin) - the muted
group_colors <- setNames(RColorBrewer::brewer.pal(max(3, length(stream_groups)), "Paired")[seq_along(stream_groups)], names(stream_groups))
```

## Map showing all flow gages

```{r}
gage_map_data <- flow_gage |>
  mutate(group = basin_label(huc8)) |>
  filter(!is.na(latitude), !is.na(longitude))

group_pal <- colorFactor(palette = group_colors[names(stream_groups)], domain = gage_map_data$group)

leaflet(gage_map_data) |>
  addTiles() |>
  addCircleMarkers(
    lng = ~longitude, lat = ~latitude,
    color = ~group_pal(group),
    radius = 5, stroke = FALSE, fillOpacity = 0.8,
    label = ~paste0(gage_name, " (", gage_id, ")")
  ) |>
  addLegend(position = "bottomright", pal = group_pal, values = ~group, title = "HUC8 sub-basin")
```

## Flow summary by sub-basin {.tabset .tabset-fade .tabset-pills}

```{r, include=FALSE}
# shared helper used by every tab below: filter flow_data/flow_gage down to one
# HUC8 sub-basin (by its gage_ids), plot date coverage per gage, and summarize
# the date range per gage
flow_group_gages <- function(huc8_code) {
  flow_gage |> filter(huc8 == huc8_code)
}

flow_group_data <- function(huc8_code) {
  flow_data |> filter(gage_id %in% flow_group_gages(huc8_code)$gage_id)
}

plot_date_coverage <- function(df) {
  df |> ggplot(aes(x = date, y = value, color = gage_name)) +
    geom_line() +
    scale_color_manual(values = colors_full) +
    facet_wrap(~gage_name) +
    theme_minimal() +
    theme(legend.position = "none", axis.text.x = element_text(angle = 90, hjust = 1)) +
    labs(title = "Flow vs Date by Gage", x = "Date", y = "Flow (cfs)")
}

summarize_date_range <- function(df) {
  df |>
    group_by(gage_name, gage_id) |>
    summarize(
      min_date = min(date),
      max_date = max(date),
      record_count = n(),
      .groups = "drop"
    ) |>
    arrange(gage_name)
}
```

```{r, results = "asis"}
for (group in names(stream_groups)) {
  huc8 <- stream_groups[[group]]

  cat("### ", group, "\n\n")
  cat("\n\n#### Gages\n\n")
  print(kable(flow_group_gages(huc8) |> select(location, gage_name, gage_id, agency, river_mile)))

  cat("\n\n#### Date coverage\n\n")
  print(plot_date_coverage(flow_group_data(huc8)))

  cat("\n\n#### Date range summary\n\n")
  print(kable(summarize_date_range(flow_group_data(huc8))))

  cat("\n\n")
}
```
