---
title: "Upper Klamath Lake - Water Temperature"
format: 
  html:
    embed-resources: true
---

```{r, include = F}
colors_full <-  c("#9A8822", "#F5CDB4", "#F8AFA8", "#FDDDA0", "#74A089", #Royal 2
                  "#899DA4", "#C93312", "#DC863B", # royal 1 (- 3)
                  "#F1BB7B", "#FD6467", "#5B1A18", # Grand Budapest 1 (-4)
                  "#D8B70A", "#02401B", "#A2A475", # Cavalcanti 1
                  "#E6A0C4", "#C6CDF7", "#D8A499", "#7294D4", #Grand Budapest 2
                  "#9986A5", "#EAD3BF", "#AA9486", "#B6854D", "#798E87" # Isle of dogs 2 altered slightly
)
```

The goal of this document is to provide a script for pulling Upper Klamath Lake
water temperature data and visualize the data.

Please install `klamathWaterData` 

`#install.packages("remotes")`

`remotes::install_github("Klamath-SDM/klamathWaterData")`

**Upper Klamath Lake USGS Gage Numbers**

- RATTLESNAKE POINT - RPT 422042121513100
- UPPER KLAMATH LAKE AT HOWARD BAY, OR 421935121551200
- MID-TRENCH - LOWER - MDTL 422305121553800
- MID-TRENCH - UPPER - MDTU 422305121553803
- SHOALWATER BAY - SHB 422444121580400
- MID-NORTH - LOWER - MDNL 422622122004000
- MID-NORTH - UPPER - MDNU 422622122004003
- WILLIAMSON RIVER OUTLET - WMR 422719121571400
- SEVENMILE CNL AT DIKE RD BR, NR KLAMATH AGENCY, OR 11504290

# UKL temperature data

```{r}
# Define UKL gages
ukl_gage_numbers <- c("422042121513100", "421935121551200", "422305121553800",
                      "422305121553803", "422444121580400", "422622122004000",
                      "422622122004003", "422719121571400","11504290")

# Filter the water temperature dataset in klamathWaterData
ukl_temp <- klamathWaterData::temperature_data |> 
  dplyr::filter(gage_id %in% ukl_gage_numbers)
```

# Visualizations

Use the plotly figure to zoom in on specific time frames or select one location.

## Visualize mean water temperature

Mean water temperature is generally consistent across locations. Temperature is
at or below 10C in April and May and increases to 25C or above by July/August.

Plots for max and min are not shown.

- Max water temperature is generally consistent across locations though Mid Trench Upper
reaches the highest temperatures. 
- Min water temperature is generally consistent across locations and is similar to 
mean temperature.

```{r}
plot <- ukl_temp |> 
  dplyr::filter(statistic == "mean") |> 
  ggplot2::ggplot(ggplot2::aes(x = date, y = value, color = gage_name)) +
  ggplot2::geom_line() +
  ggplot2::scale_color_manual(values = colors_full) +
  ggplot2::theme_bw() +
  ggplot2::labs(x = "",
       y = "Water temperature (C) - mean",
       color = "USGS Gage Name")

plotly::ggplotly(plot)
```
