---
title: "Gerber Reservoir Water Quality Data Exploration"
author: "Badhia Yunes Katz"
date: "September 2026"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 9, fig.height = 4.5)
library(tidyverse)
library(readxl)
library(janitor)
library(lubridate)
library(leaflet)
library(knitr)
```

This report explores the Gerber Reservoir water quality workbook
(`data-raw/skinner-wq-data/Gerber WQ.xlsx`) to understand its temporal coverage, data gaps and
collection points, and how it could be formatted like the other water quality objects in klamathWaterData.

The workbook has five tabs:

| Tab | Content |
|---|---|
| Gerber net pen locations | The two monitoring sites (net pens) |
| Gerber_sondes | Hourly YSI / HOBO sonde data (temperature, pH, DO, conductivity, etc.) |
| Gerber chla (SFEI) | Daily satellite-derived chlorophyll-a summaries |
| Gerber cyano | Discrete cyanobacteria / cyanotoxin samples (7 rows) |
| Cyano table for email report | A one-row formatted copy of the latest cyano sample |

```{r read}
f <- "data-raw/skinner-wq-data/Gerber WQ.xlsx"
excel_date <- function(x) as.Date(as.numeric(x), origin = "1899-12-30")

sites <- read_excel(f, sheet = "Gerber net pen locations", .name_repair = "minimal") |>
  clean_names() |>
  separate(coordinates, c("latitude", "longitude"), sep = ",", convert = TRUE)

# "NA" is stored as literal text in many cells, so read as text first
sonde_cols <- c("temp", "p_h", "do_conc", "do_sat", "sp_cond", "sal", "tds", "depth", "bat", "cab")

sonde <- read_excel(f, sheet = "Gerber_sondes", col_types = "text", .name_repair = "minimal") |>
  clean_names() |>
  mutate(across(everything(), ~ na_if(.x, "NA")),
         date = excel_date(date),
         across(all_of(sonde_cols), as.numeric)) |>
  filter(!is.na(site))   # drops one stray row with only a date (2026-10-01)

chla <- read_excel(f, sheet = "Gerber chla (SFEI)", col_types = "text", .name_repair = "minimal") |>
  clean_names() |>
  mutate(across(everything(), ~ na_if(.x, "NA")),
         date = excel_date(date),
         across(c(pixelcount, chla_mean, chla_med, chla_90per), as.numeric))

cyano <- read_excel(f, sheet = "Gerber cyano", .name_repair = "minimal") |>
  clean_names() |>
  mutate(date = excel_date(date))
```

# 1. Collection points

```{r sites}
site_tbl <- sonde |>
  group_by(site) |>
  summarise(n_readings = n(), first = min(date), last = max(date)) |>
  left_join(sites, by = "site") |>
  select(site, site_description, latitude, longitude, n_readings, first, last)

kable(site_tbl)
```

- There are **two sites**, both floating net pens: `Gerber_NP` (south) and `Gerber_NPN` (north).
- `Gerber_NP` has the older record (2021-2025); `Gerber_NPN` only starts in 2026, so the two are **not sampled concurrently**.
- The cyano tab lists slightly different coordinates for `Gerber_NP` (42.18273, -121.10111) than the locations tab; the locations tab is used here.
- The workbook contains no sonde or sample data for 2022.

```{r map}
site_tbl |>
  leaflet() |>
  addProviderTiles("Esri.WorldImagery") |>
  addCircleMarkers(~longitude, ~latitude, radius = 8, color = "white", weight = 1,
                   fillColor = "steelblue", fillOpacity = 0.9,
                   label = ~paste0(site, " - ", site_description),
                   popup = ~paste0("<b>", site, "</b><br>", site_description,
                                   "<br>", format(n_readings, big.mark = ","), " readings"),
                   labelOptions = labelOptions(noHide = TRUE, direction = "right"))
```

# 2. Sonde data (hourly)

## 2.1 Temporal coverage

```{r deployments}
deployments <- sonde |>
  group_by(site, sonde, year) |>
  summarise(first = min(date), last = max(date),
            days_with_data = n_distinct(date),
            span_days = as.numeric(max(date) - min(date)) + 1,
            .groups = "drop") |>
  mutate(missing_days_in_span = span_days - days_with_data)

kable(deployments, caption = "Deployment periods")
```

```{r sonde-timeline}
sonde |>
  distinct(site, sonde, date) |>
  mutate(label = paste(site, sonde, sep = " / ")) |>
  ggplot(aes(date, label)) +
  geom_point(shape = "|", size = 5, colour = "steelblue") +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  labs(x = NULL, y = NULL, title = "Days with sonde data") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

- Data are **seasonal deployments** (roughly Apr/May to Sep/Oct), not year-round. Winter is never sampled.
- Readings are hourly (a few 2-5 hour breaks). The 2026 `Gerber_NP` HOBO record (28 May - 25 Jun) is the only non-YSI data.
- The only real gap within a deployment year is 2024 at `Gerber_NP`: no July data, a 50-day break before 14 Aug (49 missing days).

```{r monthly-heat}
sonde |>
  mutate(year = year(date), month = month(date, label = TRUE)) |>
  count(year, month) |>
  ggplot(aes(month, factor(year), fill = n)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = n), size = 3) +
  scale_fill_viridis_c(name = "Readings") +
  labs(x = NULL, y = NULL, title = "Hourly readings per month (a full month is ~720-744)") +
  theme_minimal()
```

## 2.2 Parameter completeness

```{r completeness}
sonde |>
  mutate(deployment = paste(site, sonde, year, sep = " / ")) |>
  select(deployment, all_of(sonde_cols)) |>
  pivot_longer(-deployment, names_to = "parameter", values_to = "value") |>
  group_by(deployment, parameter) |>
  summarise(pct = 100 * mean(!is.na(value)), .groups = "drop") |>
  ggplot(aes(parameter, deployment, fill = pct)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = round(pct)), size = 3) +
  scale_fill_viridis_c(limits = c(0, 100), name = "% present") +
  labs(x = NULL, y = NULL, title = "% of readings with a value") +
  theme_minimal()
```

- Temperature and DO concentration are complete in every deployment.
- The **HOBO** record only has temperature and DO concentration (no pH, conductivity or depth).
- Salinity, TDS, "bat" and "cab" (not defined in the workbook - probably battery and cable/wiper) are only partly filled and are not water quality variables of interest.

## 2.3 Value ranges

```{r ranges}
sonde |>
  select(all_of(c("temp", "p_h", "do_conc", "do_sat", "sp_cond", "depth"))) |>
  pivot_longer(everything(), names_to = "parameter", values_to = "value") |>
  group_by(parameter) |>
  summarise(n = sum(!is.na(value)), min = min(value, na.rm = TRUE),
            median = median(value, na.rm = TRUE), max = max(value, na.rm = TRUE)) |>
  kable(digits = 2)
```

Some values look extreme for a reservoir (pH 5.7-10.1, DO saturation up to 222%, DO down to 1.7 mg/L). These
may be real bloom or sensor-fouling effects and would need a QC pass before publishing.

```{r ts, fig.height = 7}
sonde |>
  mutate(datetime = as.POSIXct(date) + as.numeric(time) * 86400,
         doy_date = as.Date(paste0("2000-", format(date, "%m-%d")))) |>
  select(datetime, doy_date, year, site, temp, do_conc, p_h) |>
  pivot_longer(c(temp, do_conc, p_h), names_to = "parameter", values_to = "value") |>
  drop_na(value) |>
  ggplot(aes(doy_date, value, colour = year)) +
  geom_point(size = 0.2, alpha = 0.5) +
  facet_wrap(~parameter, scales = "free_y", ncol = 1) +
  scale_x_date(date_labels = "%b") +
  scale_colour_brewer(palette = "Dark2") +
  guides(colour = guide_legend(override.aes = list(size = 4, alpha = 1))) +
  labs(x = NULL, y = NULL, colour = "Year", title = "Hourly sonde readings by day of year") +
  theme_minimal() +
  theme(legend.position = "top", legend.text = element_text(size = 11))
```

# 3. Satellite chlorophyll-a (SFEI)

```{r chla}
chla |>
  mutate(has_data = !is.na(chla_mean)) |>
  group_by(year) |>
  summarise(days = n(), days_with_data = sum(has_data),
            pct_with_data = round(100 * mean(has_data)),
            max_chla_mean = max(chla_mean, na.rm = TRUE)) |>
  kable(caption = "Daily satellite chlorophyll-a records")
```

- One record per day from **`r min(chla$date)`** to **`r max(chla$date)`**, which is the longest record in the workbook and covers years with no sonde data (2022).
- About **40% of days have no valid pixels** (`pixelcount` = 0, presumably cloud cover); `chla_*` are blank on those days.
- Values are whole-reservoir summaries (mean, median, 90th percentile), not tied to a net pen. Units are not stated (assumed ug/L).
- The maximum daily mean is an identical 134.6 in 2021, 2022 and 2023, which looks like a possible artifact worth checking with SFEI.

```{r chla-plot}
chla |>
  ggplot(aes(date, chla_mean)) +
  geom_point(size = 0.6, alpha = 0.6, colour = "darkgreen", na.rm = TRUE) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(x = NULL, y = "Mean chl-a", title = "Satellite chlorophyll-a (daily mean)") +
  theme_minimal()
```

# 4. Cyanobacteria samples

```{r cyano}
cyano |>
  select(date, site, atx, cyl, mc, stx, taxa) |>
  kable(caption = "Cyanotoxin results (ATX = anatoxin-a, CYL = cylindrospermopsin, MC = microcystin, STX = saxitoxin; presumably ug/L)")
```

- Only **7 discrete samples**, in summer 2024 (1), 2025 (3) and 2026 (3). Too few for a time series.
- Toxins were non-detect (0) in all but one sample: microcystin of 3.43 on 2025-09-08.
- Bloom observations and taxa are in free text (`taxa`, `notes`).
- The "Cyano table for email report" tab is a formatted copy of the 2026-09-01 sample, with `ND` for non-detect.

# 5. Formatting for klamathWaterData

Existing continuous-data objects (`temperature_data`, `do_data`, `ph_data`) store **daily** `mean` / `minimum` /
`maximum` values with columns `location, gage_name, gage_id, variable_name, value, unit, statistic, date`.
The hourly sonde data can be summarised to that format. Draft conversion (days with at least 20 hourly readings only):

```{r package-format}
param_map <- tribble(
  ~col,      ~variable_name,      ~unit,
  "temp",    "temperature",       "celsius",
  "do_conc", "dissolved oxygen",  "mg/L",
  "p_h",     "pH",                "std unit"
)

gerber_wq_data <- sonde |>
  select(site, date, all_of(param_map$col)) |>
  pivot_longer(all_of(param_map$col), names_to = "col", values_to = "value") |>
  drop_na(value) |>
  group_by(site, date, col) |>
  summarise(n = n(), mean = mean(value), minimum = min(value), maximum = max(value), .groups = "drop") |>
  filter(n >= 20) |>
  pivot_longer(c(mean, minimum, maximum), names_to = "statistic", values_to = "value") |>
  left_join(param_map, by = "col") |>
  mutate(location  = "gerber reservoir",
         gage_name = paste("gerber reservoir", if_else(site == "Gerber_NP", "net pen south", "net pen north")),
         gage_id   = paste0("GERBER-", str_remove(site, "Gerber_"))) |>
  select(location, gage_name, gage_id, variable_name, value, unit, statistic, date) |>
  as.data.frame()

gerber_wq_gage <- site_tbl |>
  transmute(gage_name = paste("gerber reservoir", if_else(site == "Gerber_NP", "net pen south", "net pen north")),
            gage_id = paste0("GERBER-", str_remove(site, "Gerber_")),
            agency = NA_character_, latitude, longitude,
            river_mile = NA_real_, huc8 = NA_character_, location = "gerber reservoir")

count(gerber_wq_data, gage_id, variable_name, statistic, name = "n_days") |>
  pivot_wider(names_from = statistic, values_from = n_days) |>
  mutate(existing_object = case_when(variable_name == "temperature" ~ "temperature_data",
                                     variable_name == "dissolved oxygen" ~ "do_data",
                                     variable_name == "pH" ~ "ph_data")) |>
  kable(caption = "Daily records per site and variable")
```

# 6. Summary and open questions

**Coverage**

- Hourly sonde data at two net pens: `Gerber_NP` in 2021 and 2023-2025, `Gerber_NPN` in 2026 (plus a short 2026 HOBO at `Gerber_NP`).
- Deployments are seasonal (about Apr-Oct); 2022 is missing entirely and 2024 has no July data.
- Satellite chl-a is daily since Jan 2021 but is blank on ~40% of days; cyanotoxin samples are only 7 rows.

**Decisions needed before adding to the package**

1. **Sonde summarisation** - daily mean/min/max (as drafted, with a 20-reading minimum) or keep hourly values?
2. **Which variables** - temperature, DO and pH map to existing objects; conductivity, DO saturation and depth would need new objects.
3. **Chlorophyll-a and cyanotoxins** - the package has no chl-a or HAB objects yet. Add new objects, or leave out?
4. **QC** - pH and DO extremes (see section 2.3) need review; confirm whether the provider has already QC'd the data.
5. **HOBO vs YSI** - keep the 2026 HOBO record (temperature and DO only) alongside YSI data?
6. **`location` / `agency` / `huc8`** - "gerber reservoir" is a placeholder; confirm data owner (agency) and HUC8.
7. **Units and definitions** - chl-a units, `bat` / `cab` meaning and cyanotoxin units are not documented in the workbook.
8. **Stray row** - one row with only a date (2026-10-01) was dropped.
