---
title: "HAB data Exploration "
author: "Badhia Yunes Katz"
date: "August 2025"
# output: rmarkdown::github_document
---


```{r include=FALSE}
library(tidyverse)
library(dataRetrieval)
library(leaflet)
library(janitor)
library(dplyr)
library(ggplot2)
library(DT)         
library(tidyr)
library(scales)
library(leaflet)
library(htmltools)
```


This report is has an overview of harmful algal bloom (HAB) data and practice in the Klamath Basin (southern Oregon & northern California). It covers:

- What HAB parameters matter and the thresholds used in health guidance (with links to sources)
- What data exists in the Water Quality Portal (WQX) for the basin
- Data pull those data for the last ~10 years, flag likely HAB events, and map them
- Caveats for interpreting results


# 2. Basin context

The Klamath system includes Upper Klamath Lake, the Klamath River, and major tributaries. 
HABs occur as:

- Planktonic blooms (e.g., Microcystis) → often produce microcystins (toxin)
- Benthic cyanobacterial mats (e.g., Microcoleus) → can produce anatoxin-a, especially in river reaches

**Resources**  
- [Oregon Health Authority (OHA) HAB program (advisories, criteria, program logic)](https://www.oregon.gov/oha/PH/HEALTHYENVIRONMENTS/RECREATION/HARMFULALGAEBLOOMS/Pages/index.aspx)
- [USGS Klamath studies overview](https://www.usgs.gov/centers/oregon-water-science-center/science/upper-klamath-basin-studies-and-data-collection)


# 3. HAB parameters (thresholds & sources)

These are the most decision-relevant parameters for recreational waters. We use them to classify toxic events and bloom evidence.

| Parameter | Threshold (recreation context) | Why it matters | Sources (working links) |
|---|---:|---|---|
| **Microcystin** | **≥ 8 µg/L** | Most common cyanotoxin; core basis for advisories | [EPA 2019 Criteria (PDF)](https://www.epa.gov/sites/default/files/2019-05/documents/hh-rec-criteria-habs-document-2019.pdf) ; [OHA 2024 Guidance (PDF)](https://www.oregon.gov/oha/PH/HEALTHYENVIRONMENTS/RECREATION/HARMFULALGAEBLOOMS/Documents/Advisory%20Guidelines%20for%20Harmful%20Cyanobacteria%20Blooms%20in%20Recreational%20Waters.2024_Final.pdf) |
| **Cylindrospermopsin** | **≥ 15 µg/L** | Cyanotoxin; used in EPA/OHA frameworks | [EPA 2019 Criteria (PDF)](https://www.epa.gov/sites/default/files/2019-05/documents/hh-rec-criteria-habs-document-2019.pdf) |
| **Anatoxin-a** | **≥ 15 µg/L** | Neurotoxin; common in benthic mats | [OHA 2024 Guidance (PDF)][https://www.oregon.gov/oha/PH/HEALTHYENVIRONMENTS/RECREATION/HARMFULALGAEBLOOMS/Documents/Advisory%20Guidelines%20for%20Harmful%20Cyanobacteria%20Blooms%20in%20Recreational%20Waters.2024_Final.pdf] |
| **Saxitoxin** | **≥ 8 µg/L** | Neurotoxin; advisory value | OHA 2024 Guidance (PDF): link above |
| **Cyanobacteria (cells)** | **≥ 100,000 cells/mL** | WHO alert-level trigger; risk management | [WHO 2nd ed. *Toxic Cyanobacteria in Water* (landing)](https://www.who.int/publications/m/item/toxic-cyanobacteria-in-water---second-edition); [Full open PDF](https://library.oapen.org/bitstream/20.500.12657/47047/1/9781000262025.pdf) |
| **Chlorophyll a** | **≈ 20–50 µg/L** when cyano-dominated; **use ≥ 30 µg/L** for conservative screening | Biomass proxy; supports bloom identification | WHO 2nd ed., Ch. 5 (open PDF): link above |
| **Phycocyanin** | **Site-specific (no universal cut-point)**; use **local 90th percentile** as screening | Cyanobacteria pigment (sensor reading); good early warning when calibrated to cells/toxins | [EPA 2019 Monitoring Guidance (PDF)](https://www.epa.gov/sites/default/files/2019-09/documents/recommend-cyano-rec-water-2019-update.pdf); Review: [Almuhtaram et al., 2021 (overview of sensors and variability)](https://www.sciencedirect.com/science/article/pii/S1470160X21011079) |


Benthic vs planktonic: In the Klamath River, benthic mats can drive anatoxin-a even when the water column seems clear—plan monitoring accordingly (see e.g., Genzoli et al., 2024, Freshwater Science).

---

# 4. What data is available

- **Water Quality Portal (WQX/NWIS/STORET):** In-situ measurements reported by agencies/tribes/utilities. We will pull the 7 parameters above for the Klamath Basin. https://www.waterqualitydata.us/
- **OHA HAB advisories (issue/lift dates):** Official public-health decisions for Oregon waters. https://www.oregon.gov/oha/PH/HEALTHYENVIRONMENTS/RECREATION/HARMFULALGAEBLOOMS/Pages/Blue-GreenAlgaeAdvisories.aspx
- **KBMP (Klamath Basin Monitoring Program) Blue-Green Algae Tracker:** Reach-level flags where thresholds are exceeded; useful spatial context. https://kbmp.net/maps-and-data
- **CyAN (EPA/NOAA/USGS) satellite index:** Weekly cyanobacteria index for large waters (e.g., Upper Klamath Lake). Public viewer: https://experience.arcgis.com/experience/01f798892ecb48d8ac99ef9b4ebe37c8

This markdown focuses on pulling data from WQX.


# 5. Toxic “events”

- **Toxic HAB event (primary evidence):** any sample with **Microcystin ≥ 8 µg/L** (or other cyanotoxin ≥ its threshold).
- **Bloom evidence (supporting):** **Cyanobacteria ≥ 100,000 cells/mL**, **Chlorophyll-a ≥ 30 µg/L** (screening), or **Phycocyanin ≥ 90th percentile** locally.
- **Interpretation tip:** If toxins are absent but **benthic mats** are present, consider **anatoxin-a** detection and site context.

---

# 6. Pulling WQX data (last ~10 years for Klamath Basin (HUC2 = 180102))


The workflow:

1. Query WQP for relevant parameters  
2. Summarize maximum values per year and site  
3. Identify likely bloom events  
4. Map bloom locations and provide a summary table  


```{r include=FALSE}
# Define Klamath Basin HUC8 codes
huc_code <- "180102"

hab_characteristics <- c("Phycocyanin", "Chlorophyll a","Microcystin","Cyanobacteria", "Anatoxin-A", "Saxitoxin", "Cylindrospermopsin")

# Set date range
start_date <- "2014-01-01"
end_date <- "2025-12-31"
```

```{r include=FALSE}
# Pull result data
wqx_results <- map(hab_characteristics, function(cn) {
  message("Querying WQP for: ", cn)
  tryCatch(
    readWQPdata(
      huc = huc_code,
      characteristicName = cn,
      startDate = start_date,
      endDate   = end_date,
      sampleMedia = "Water"
    ),
    error = function(e) { warning("WQP error for ", cn, ": ", e$message); NULL }
  )
}) |> compact()

stopifnot(length(wqx_results) > 0)    # basic guard

harmonize_wqx <- function(df) {
  df <- janitor::clean_names(df)

  # Ensure key columns exist
  needed <- c(
    "result_measure_value",
    "result_measure_measure_unit_code",
    "activity_start_date",
    "monitoring_location_identifier",
    "monitoring_location_name",
    "characteristic_name",
    "latitude_measure",
    "longitude_measure"
  )
  for (nm in needed) if (!nm %in% names(df)) df[[nm]] <- NA_character_

  # Force consistent types BEFORE binding; parse later
  df |>
    mutate(
      result_measure_value = as.character(result_measure_value),
      result_measure_measure_unit_code = as.character(result_measure_measure_unit_code),
      activity_start_date = as.character(activity_start_date),
      latitude_measure = as.character(latitude_measure),
      longitude_measure = as.character(longitude_measure),
      characteristic_name = as.character(characteristic_name),
      monitoring_location_identifier = as.character(monitoring_location_identifier),
      monitoring_location_name = as.character(monitoring_location_name)
    )
}

# Apply harmonization to each chunk, then bind
all_results <- wqx_results |>
  purrr::map(harmonize_wqx) |>
  dplyr::bind_rows()

# Now safely coerce to usable types
all_results <- all_results |>
  mutate(
    date      = as.Date(activity_start_date),
    value_num = suppressWarnings(readr::parse_number(result_measure_value)),
    latitude_measure  = suppressWarnings(as.numeric(latitude_measure)),
    longitude_measure = suppressWarnings(as.numeric(longitude_measure)))

# Pull site metadata to get names and coordinates
site_info <- whatWQPsites(huc = huc_code) |> clean_names()

# Join & prepare fields
dat <- all_results |>
  left_join(site_info, by = "monitoring_location_identifier")

glimpse(dat)
```

```{r echo=FALSE}
# Toxins (EPA 2019; OHA 2024)
thr_microcystin        <- 8    # µg/L
thr_cylindrospermopsin <- 15   # µg/L
thr_anatoxin_a         <- 15   # µg/L
thr_saxitoxin          <- 8    # µg/L

# Supporting indicators (WHO/literature)
thr_cyano_cells <- 100000      # cells/mL
thr_chla        <- 30          # µg/L (conservative screening)

# Phycocyanin "elevated": local 90th percentile (screening)
pc90 <- dat |>
  filter(characteristic_name == "Phycocyanin") |>
  summarise(pc90 = quantile(value_num, 0.90, na.rm = TRUE)) |>
  pull(pc90)
if (length(pc90) == 0 || is.na(pc90)) pc90 <- Inf

unit_safe <- function(x) ifelse(is.na(x) | x=="", "", x)

# Compute a Phycocyanin 90th percentile (across available pc data)
pc90 <- dat |>
  dplyr::filter(characteristic_name == "Phycocyanin", is.finite(value_num)) |>
  dplyr::summarise(pc90 = stats::quantile(value_num, 0.90, na.rm = TRUE)) |>
  dplyr::pull(pc90)
if (length(pc90) == 0 || is.na(pc90)) pc90 <- Inf

# Helper predicates to guard by unit where appropriate
is_ugL <- function(u) grepl("^\\s*ug/?L\\s*$", u, ignore.case = TRUE)  # µg/L variants
is_cells_ml <- function(u) grepl("cells\\s*/\\s*mL", u, ignore.case = TRUE)


hab_events <- dat |>
  dplyr::mutate(toxic_event = dplyr::case_when(
    characteristic_name == "Microcystin" & is_ugL(result_measure_measure_unit_code) & value_num >= thr_microcystin ~ TRUE,
    characteristic_name == "Cylindrospermopsin" & is_ugL(result_measure_measure_unit_code) & value_num >= thr_cylindrospermopsin ~ TRUE,
    characteristic_name == "Anatoxin-a" & is_ugL(result_measure_measure_unit_code) & value_num >= thr_anatoxin_a ~ TRUE,
    characteristic_name == "Saxitoxin" & is_ugL(result_measure_measure_unit_code) & value_num >= thr_saxitoxin ~ TRUE,
    TRUE ~ FALSE),
    # Supporting indicators 
    bloom_support = dplyr::case_when(
      characteristic_name == "Cyanobacteria" & is_cells_ml(result_measure_measure_unit_code) & value_num >= thr_cyano_cells ~
        TRUE,
      characteristic_name == "Chlorophyll a" & is_ugL(result_measure_measure_unit_code) & value_num >= thr_chla ~ TRUE,
      characteristic_name == "Phycocyanin" & is.finite(value_num) & value_num >= pc90 ~ TRUE,
      TRUE ~ FALSE),
    event = toxic_event | bloom_support) |>
  dplyr::filter(event) |> 
  dplyr::select(
    date, monitoring_location_identifier, 
    monitoring_location_name.y,
    characteristic_name, value_num, result_measure_measure_unit_code,
    latitude_measure.y, longitude_measure.y, toxic_event, bloom_support)

head(hab_events, 10)
```

Yearly context: per-site maxima

```{r echo=FALSE}
summary_table <- dat |> 
  mutate(year = year(date)) |> 
  group_by(monitoring_location_identifier, monitoring_location_name.y,
           characteristic_name, year) |> 
  summarise(samples = dplyr::n(),
            max_value = suppressWarnings(max(value_num, na.rm = TRUE)),
            units = dplyr::first(result_measure_measure_unit_code),
            .groups   = "drop") |> 
  arrange(desc(max_value))

head(summary_table, 10)
```


```{r echo=FALSE}
ggplot(summary_table, aes(x = year, y = max_value)) +
  geom_point() +
  facet_wrap(~ characteristic_name, scales = "free_y") +
  labs(
    title = "Maximum HAB-Related Values by Year (per parameter)",
    x = "Year", y = "Max value (units vary by parameter)")
```


```{r eval=FALSE, include=FALSE}
pal <- colorFactor(
  palette = c("darkred","darkgreen","blue","orange","purple","brown","goldenrod"),
  domain  = hab_events$characteristic_name)

hab_events <- hab_events |> 
  mutate(latitude_measure.y = as.numeric(latitude_measure.y),
         longitude_measure.y = as.numeric(longitude_measure.y))

leaflet(hab_events) |> 
  addTiles() |> 
  addCircleMarkers(lng = ~longitude_measure.y,  # small jitter
                   lat = ~latitude_measure.y,
    color = ~pal(characteristic_name),
    label = ~paste0(sprintf(
      "<b>%s</b><br><b>Date:</b> %s<br><b>%s:</b> %s %s",
      htmlEscape(monitoring_location_name.y),
      format(date, "%Y-%m-%d"),
      htmlEscape(characteristic_name),
      number(value_num, accuracy = 0.01),
      htmlEscape(unit_safe(result_measure_measure_unit_code)))),
    radius = 6, fillOpacity = 0.8) |> 
  addLegend("bottomright", pal = pal, values = ~characteristic_name, title = "HAB Parameter")

```



```{r echo=FALSE}
# color by risk category
hab_events <- hab_events |> 
  mutate(latitude_measure.y  = suppressWarnings(as.numeric(latitude_measure.y)),  
         longitude_measure.y = suppressWarnings(as.numeric(longitude_measure.y)),
         event_type = case_when(
           isTRUE(toxic_event) ~ "Toxic event", # derive event category from flags
           !isTRUE(toxic_event) & isTRUE(bloom_support) ~ "Bloom-support",
           TRUE ~ "Other")) |> 
  # keep only rows with valid coordinates
  filter(is.finite(latitude_measure.y), is.finite(longitude_measure.y))

etype_pal <- colorFactor(
  palette = c("Toxic event" = "red", "Bloom-support" = "goldenrod", "Other" = "gray50"),
  domain  = c("Toxic event", "Bloom-support", "Other"))

hab_events <- hab_events |> 
  mutate(popup_html = sprintf(
    "<b>%s</b><br/><b>Date:</b> %s<br/><b>Parameter:</b> %s<br/><b>Value:</b> %s %s<br/><b>Type:</b> %s",
    htmlEscape(monitoring_location_name.y),
    format(date, "%Y-%m-%d"),
    htmlEscape(characteristic_name),
    number(value_num, accuracy = 0.01),
    htmlEscape(unit_safe(result_measure_measure_unit_code)),
    event_type))


leaflet(hab_events) |> 
  addTiles() |> 
  addCircleMarkers(lng = ~longitude_measure.y,
                   lat   = ~latitude_measure.y,
                   color = ~etype_pal(event_type),
                   radius = 6, fillOpacity = 0.85, stroke = TRUE, weight = 1,
                   popup  = ~popup_html) |> 
  addLegend("bottomright",
            pal = etype_pal,
            values = ~event_type,
            title = "Event type")

```
