---
title: "Elevation Crosswalk: USBR Hydromet/USGS Elevations & WET Surface Water Extent"
author: "Maddee Wiggins"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

```{r libraries}
library(tidyverse)
library(sf)
library(ggplot2)
library(scales)
library(patchwork)
library(knitr)
library(leaflet)
```

# Overview

This document crosswalks water-surface-**elevation** stations from `teacup_diagram_data`
(https://www.usbr.gov/pn/hydromet/klamath/teacup.html) to Klamath Basin wildlife refuges
(`rivermile::klamath_refuges`) and lakes/reservoirs (`rivermile::klamath_lakes`), then joins
each matched station's monthly water level to the processed WET (Intermountain West Joint
Venture (IWJV)) monthly surface water extent data (`wet_surface_water`) for that area. The goal
is a combined monthly view — station water level alongside wetland/lake surface water extent —
for areas with a relevant elevation gauge.

The area/station crosswalk is a hand-curated lookup (Section 4); only stations that are
actually on or clearly hydrologically tied to an area are assigned to it. `wet_surface_water`/
`wet_hydroperiod` key on a `name` column with a `boundary_type` of `"refuge"` or `"lake"` — see
`data-raw/processing-scripts/wet-process-data.R` for how those ids are built. **Bear Valley,
Klamath Marsh, Lower Klamath/Sheepy refuges, and Lower Klamath Lake** have no station assigned
and are excluded from the combined analysis.

---

# 1. Data processing notes

1. **Surface water extent** (`wet_surface_water`/`wet_hydroperiod`) is calculated within a
   **0.5-mile buffer** around each refuge/lake polygon (`rivermile::klamath_refuges` /
   `rivermile::klamath_lakes`), not the raw polygon boundary itself — see
   `data-raw/processing-scripts/wet-process-data.R`.
2. **Elevation** is currently crosswalked from `teacup_diagram_data` as the default source
   (Section 4). This is a hand-curated, on-site-only crosswalk, not exhaustive — additional
   gages at **Clear Lake** and **Upper Klamath Lake** will be explored as candidate additions
   or alternatives.
3. See [Section 4.1](#maps-refuges-lakes-with-elevation-stations) for a map showing all
   stations against the refuge/lake boundaries.

---

# 2. Load data

```{r load-data}
data("wet_surface_water",    package = "klamathWaterData")
data("teacup_diagram_data",  package = "klamathWaterData")

refuges <- rivermile::klamath_refuges
lakes <- rivermile::klamath_lakes 
 
cat("wet_surface_water:  ", nrow(wet_surface_water),   "rows\n")
cat("teacup_diagram_data:", nrow(teacup_diagram_data), "rows\n")
cat("refuges:            ", nrow(refuges),             "features\n")
cat("lakes:              ", nrow(lakes),               "features\n")

kable(st_drop_geometry(refuges), caption = "rivermile::klamath_refuges")
kable(st_drop_geometry(lakes),   caption = "rivermile::klamath_lakes")
```

---

# 3. Elevation stations in `teacup_diagram_data`

`teacup_diagram_data$measure_type` is one of `elevation`, `storage`, or `flow`. We keep only
`elevation` — reservoir/lake water-surface elevation and gauge height, which is what can
plausibly track an area's wetland/lake extent. `lat`/`long` are already included in this
dataset (sourced from the USBR RISE API for USBR Hydromet sites, and
`dataRetrieval::readNWISsite()` for USGS sites — see
`data-raw/data-pull/teacup-diagram-data-pull.R`).

```{r elevation-stations}
elev_stations <- teacup_diagram_data |>
  filter(measure_type == "elevation") |>
  distinct(source, site, label, lat, long)

stations_sf <- elev_stations |>
  st_as_sf(coords = c("long", "lat"), crs = 4326, remove = FALSE)

kable(elev_stations, caption = "Elevation stations available in teacup_diagram_data")
```

---

# 4. Spatial crosswalk of elevation sites with refuges and lakes

`wet_surface_water`/`wet_hydroperiod` identify each area with a `name` id built in
`data-raw/processing-scripts/wet-process-data.R`: refuge ids are suffixed `_refuge` (e.g.
`clear_lake_refuge`), lake ids are not (e.g. `clear_lake_reservoir`) — the two id lookups below
mirror that script exactly so the crosswalk here lines up with the actual dataset.

```{r area-id-lookup}
refuge_id_lookup <- tribble(
  ~name,                         ~orgname,
  "upper_klamath_lake_refuge",   "upper klamath national wildlife refuge",
  "clear_lake_refuge",           "clear lake national wildlife refuge",
  "tule_lake_refuge",            "tule lake national wildlife refuge",
  "bear_valley_refuge",          "bear valley national wildlife refuge",
  "lower_klamath_sheepy_refuge", "lower klamath national wildlife refuge",
  "klamath_marsh_refuge",        "klamath marsh national wildlife refuge"
)

lake_id_lookup <- tribble(
  ~name,                   ~lake_name,
  "clear_lake_reservoir",  "clear lake reservoir",
  "gerber_reservoir",      "gerber reservoir",
  "lower_klamath_lake",    "lower klamath lake",
  "tule_lake",             "tule lake",
  "upper_klamath_lake",    "upper klamath lake"
)

refuges_named <- refuges |> inner_join(refuge_id_lookup, by = "orgname")
lakes_named   <- lakes   |> inner_join(lake_id_lookup,   by = "lake_name")
```

Each elevation station is assigned to an area only where it is genuinely on-site or the
obvious controlling structure — not just the nearest point on a map:

- **Tule Lake NWR** and **Tule Lake** (the lake itself) — `TULC` and `TULC2`, the Sump 1A and
  Sump 1B forebay gauges inside the refuge boundary.
- **Clear Lake NWR** and **Clear Lake Reservoir** — `CLK`, Clear Lake West Lobe, inside both
  boundaries.
- **Upper Klamath Lake NWR** and **Upper Klamath Lake** — `11507001`, the USGS lake-elevation
  gauge near Link River Dam.
- **Gerber Reservoir** — `GER`, the reservoir's own forebay elevation gauge (there is no
  matching refuge boundary for Gerber, only the lake).
- **Bear Valley, Klamath Marsh, Lower Klamath/Sheepy refuges, and Lower Klamath Lake** — no
  station assigned.

```{r crosswalk-table}
crosswalk <- tribble(
  ~name,                       ~site,
  "tule_lake_refuge",          "TULC",
  "tule_lake_refuge",          "TULC2",
  "tule_lake",                 "TULC",
  "tule_lake",                 "TULC2",
  "upper_klamath_lake_refuge", "11507001",
  "upper_klamath_lake",        "11507001",
  "clear_lake_refuge",         "CLK",
  "clear_lake_reservoir",      "CLK",
  "gerber_reservoir",          "GER"
)

stations_area_cross <- stations_sf |>
  left_join(crosswalk, by = "site")

stations_area_cross |>
  st_drop_geometry() |>
  kable(caption = "Area x station crosswalk (NA = elevation station not assigned to an area)")
```

## 4.1 Maps: refuges & lakes with elevation stations

```{r}
# ---- teacup_diagram_data -> sf points, one row per station ----
# teacup_diagram_data is a long daily time series (197k rows); collapse to
# one point per station (some sites have multiple measure_types, e.g. GER
# has both elevation and storage — combine those into one popup).
teacup_diagram_sf <- klamathWaterData::teacup_diagram_data |>
  filter(measure_type == "elevation") |> 
  group_by(source, site, lat, long) |>
  summarise(
    measures = paste(sort(unique(measure_type)), collapse = ", "),
    labels   = paste(unique(label), collapse = "<br>"),
    .groups  = "drop"
  ) |>
  st_as_sf(coords = c("long", "lat"), crs = 4326, remove = FALSE)

# ---- water_level_gage markers (from earlier session work) ----
gages <- klamathWaterData::water_level_gage |>
  mutate(
    latitude  = if_else(gage_id == "usbr tule lake sump 1a", 41.878620, latitude),
    longitude = if_else(gage_id == "usbr tule lake sump 1a", -121.545291, longitude),
    marker_color = case_when(
      agency == "USGS"                       ~ "steelblue",
      agency == "u.s. bureau of reclamation" ~ "darkorange"
    )
  )

# ---- refuges / lakes (lakes cleaned: Tule Lake's distant outlier part dropped — see wet-process-data.R) ----
refuges <- rivermile::klamath_refuges

# ---- map ----
leaflet() |>
  addProviderTiles(providers$CartoDB.Positron) |>
  setView(lng = -121.8, lat = 42.1, zoom = 8) |>
  addPolygons(
    data = refuges,
    color = "darkgreen", weight = 1.5, fillOpacity = 0.15,
    label = ~orgname,
    group = "Refuges"
  ) |>
  addPolygons(
    data = lakes,
    color = "darkblue", weight = 1.5, fillOpacity = 0.15,
    label = ~lake_name,
    group = "Lakes"
  ) |>
  addCircleMarkers(
    data = gages,
    lng = ~longitude, lat = ~latitude,
    radius = 7,
    color = ~marker_color,
    stroke = FALSE,
    fillOpacity = 0.85,
    label = ~gage_name,
    popup = ~paste0(
      "<b>", gage_name, "</b><br>",
      "Gage ID: ", gage_id, "<br>",
      "Agency: ", agency, "<br>",
      "Location: ", location
    ),
    group = "Water Level Gages"
  ) |>
  addCircleMarkers(
    data = teacup_diagram_sf,
    radius = 6,
    color = "purple",
    stroke = TRUE, weight = 1, opacity = 1,
    fillOpacity = 0.6,
    label = ~site,
    popup = ~paste0(
      "<b>", site, "</b><br>",
      "Source: ", source, "<br>",
      "Measures: ", measures, "<br>",
      labels
    ),
    group = "Teacup Diagram Stations"
  ) |>
  addLegend(
    position = "bottomright",
    colors = c("steelblue", "darkorange", "purple"),
    labels = c("USGS gage", "USBR gage", "Teacup diagram station"),
    title = "Legend"
  ) |>
  addLayersControl(
    overlayGroups = c("Refuges", "Lakes", "Water Level Gages", "Teacup Diagram Stations"),
    options = layersControlOptions(collapsed = FALSE)
  )


```

<!-- Split into two faceted maps — one panel per refuge, one panel per lake — each zoomed to its -->
<!-- own boundary (`scales = "free"`) with the elevation stations overlaid for context. -->

<!-- ```{r crosswalk-map-refuges, fig.height=8} -->
<!-- stations_for_map <- stations_area_cross |> distinct(site, .keep_all = TRUE) |> filter(site != "MAL") -->

<!-- ggplot() + -->
<!--   geom_sf(data = refuges_named, aes(fill = orgname), alpha = 0.5, color = "grey30") + -->
<!--   geom_sf(data = stations_for_map, color = "black", size = 2) + -->
<!--   geom_sf_text(data = stations_for_map, aes(label = site), nudge_y = 0.03, size = 3) + -->
<!--   #facet_wrap(~orgname, scales = "free") + -->
<!--   scale_fill_brewer(palette = "Set3", name = "Refuge", guide = "none") + -->
<!--   labs( -->
<!--     title = "Klamath Refuges & Teacup Diagram Elevation Stations", -->
<!--     x = NULL, y = NULL -->
<!--   ) + -->
<!--   theme_minimal(base_size = 11) + -->
<!--   theme(strip.text = element_text(size = 8, face = "bold")) -->
<!-- ``` -->

<!-- ```{r crosswalk-map-lakes, fig.height=8} -->
<!-- ggplot() + -->
<!--   geom_sf(data = lakes_named, fill = "steelblue", alpha = 0.35, color = "steelblue4", linewidth = 0.6) + -->
<!--   geom_sf(data = stations_for_map, color = "black", size = 2) + -->
<!--   geom_sf_text(data = stations_for_map, aes(label = site), nudge_y = 0.03, size = 3) + -->
<!--   #facet_wrap(~lake_name, scales = "free") + -->
<!--   labs( -->
<!--     title = "Klamath Lakes & Teacup Diagram Elevation Stations", -->
<!--     x = NULL, y = NULL -->
<!--   ) + -->
<!--   theme_minimal(base_size = 11) + -->
<!--   theme(strip.text = element_text(size = 8, face = "bold")) -->
<!-- ``` -->

---

# 5. Monthly water level & WET surface water extent

`wet_surface_water` has two temporal structures: decadal composites (`is_decadal == TRUE`, one
value per calendar month averaged over a 10-year era) and individual calendar months from 2022
onward (`is_decadal == FALSE`, `era_start` is the actual year). We match each structure to the
elevation record at the same resolution: a decadal climatological monthly mean for decadal rows,
and the actual year-month mean for annual rows. Where an area has multiple stations (Tule
Lake's two sumps), their monthly values are averaged together first. Alongside the mean we also
keep the min/max elevation observed within each aggregation window, used for the range ribbon in
[Section 6.1](#ribbon). Joining directly on `name` — the same id `wet_surface_water` uses —
means both refuges and lakes fall out of one join, with no separate refuge/lake handling needed.

```{r monthly-area-xwalk}
elev_xwalk_data <- teacup_diagram_data |>
  filter(measure_type == "elevation") |>
  # many-to-many is expected: TULC/TULC2 map to both tule_lake_refuge and tule_lake,
  # 11507001 maps to both upper_klamath_lake_refuge and upper_klamath_lake, CLK maps to
  # both clear_lake_refuge and clear_lake_reservoir
  inner_join(crosswalk, by = "site", relationship = "many-to-many")

elev_monthly_decadal <- elev_xwalk_data |>
  mutate(year = year(date), month_num = month(date)) |>
  mutate(era_str = case_when(
    year >= 2005 & year <= 2014 ~ "2005-2014",
    year >= 2015 & year <= 2022 ~ "2015-2022",
    TRUE ~ NA_character_
  )) |>
  filter(!is.na(era_str)) |>
  group_by(name, era_str, month_num) |>
  summarise(elevation_mean_ft = mean(value, na.rm = TRUE),
            elevation_min_ft  = min(value, na.rm = TRUE),
            elevation_max_ft  = max(value, na.rm = TRUE), .groups = "drop") |>
  mutate(is_decadal = TRUE)

elev_monthly_annual <- elev_xwalk_data |>
  mutate(year = year(date), month_num = month(date)) |>
  filter(year >= 2022) |>
  group_by(name, era_start = year, month_num) |>
  summarise(elevation_mean_ft = mean(value, na.rm = TRUE),
            elevation_min_ft  = min(value, na.rm = TRUE),
            elevation_max_ft  = max(value, na.rm = TRUE), .groups = "drop") |>
  mutate(is_decadal = FALSE)

combined_decadal <- wet_surface_water |>
  filter(is_decadal) |>
  inner_join(elev_monthly_decadal, by = c("name", "is_decadal", "era_str", "month_num"))

combined_annual <- wet_surface_water |>
  filter(!is_decadal) |>
  inner_join(elev_monthly_annual, by = c("name", "is_decadal", "era_start", "month_num"))

combined_monthly <- bind_rows(combined_decadal, combined_annual) |>
  arrange(name, is_decadal, era_start, month_num)

plot_areas <- sort(unique(combined_monthly$name))

cat("Combined monthly rows (decadal):", nrow(combined_decadal), "\n")
cat("Combined monthly rows (annual): ", nrow(combined_annual), "\n")
cat("Areas represented:", paste(plot_areas, collapse = ", "), "\n")

kable(head(combined_monthly, 10), caption = "combined_monthly — first rows")
```

## 5.1 Package dataset: `refuge_elevation_wetland`

`combined_decadal` and `combined_annual` are working objects scoped to this document. For
downstream use in the package, they're combined (via `combined_monthly`) into a single, tidy,
renamed dataset — one row per area-month, elevation and WET columns clearly distinguished by
prefix, `boundary_type` carried through from `wet_surface_water` — and saved as
`refuge_elevation_wetland`.

```{r package-dataset}
refuge_elevation_wetland <- combined_monthly |>
  transmute(
    name,
    boundary_type,
    is_decadal,
    era_str,
    era_start,
    month_num,
    month_chr,
    elevation_mean_ft,
    elevation_min_ft,
    elevation_max_ft,
    wet_mean_val   = mean_val,
    wet_median_val = median_val,
    wet_sd_val     = sd_val,
    wet_n_pixels   = n_pixels
  ) |>
  arrange(name, is_decadal, era_start, month_num) |>
  mutate(month_chr = tolower(month_chr)) |>
  select(-era_start)

glimpse(refuge_elevation_wetland)
kable(head(refuge_elevation_wetland, 10), caption = "refuge_elevation_wetland — first rows")

usethis::use_data(refuge_elevation_wetland, overwrite = TRUE)
```

---

# 6. Combined monthly plots

Dual view of station elevation and WET mean surface water value for the areas with an assigned
station: Clear Lake NWR & Reservoir, Tule Lake NWR & the lake itself, Upper Klamath Lake NWR &
the lake itself, and Gerber Reservoir.

```{r combined-decadal-plot, fig.width=16, fig.height=8}
combined_decadal |>
  filter(name %in% plot_areas) |>
  select(name, era_str, month_num, elevation_mean_ft, mean_val) |>
  pivot_longer(c(elevation_mean_ft, mean_val), names_to = "metric", values_to = "value") |>
  mutate(metric = recode(metric,
                         elevation_mean_ft = "Water surface elevation (ft)",
                         mean_val           = "WET mean surface water value")) |>
  ggplot(aes(x = month_num, y = value, color = era_str, group = era_str)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  facet_wrap(vars(metric, name), scales = "free_y", ncol = length(plot_areas)) +
  scale_x_continuous(breaks = c(1, 4, 7, 10), labels = month.abb[c(1, 4, 7, 10)]) +
  scale_color_brewer(palette = "Set2", name = "Era") +
  labs(
    title = "Monthly Water Surface Elevation vs. WET Surface Water — Decadal Composites",
    x = "Month", y = NULL
  ) +
  theme_minimal(base_size = 9) +
  theme(legend.position = "bottom", strip.text = element_text(size = 7))
```

Annual monthly values (2022–present):

```{r combined-annual-plot, fig.width=16, fig.height=8}
combined_annual |>
  filter(name %in% plot_areas) |>
  select(name, era_start, month_num, elevation_mean_ft, mean_val) |>
  pivot_longer(c(elevation_mean_ft, mean_val), names_to = "metric", values_to = "value") |>
  mutate(metric = recode(metric,
                         elevation_mean_ft = "Mean water surface elevation (ft)",
                         mean_val           = "WET mean surface water value")) |>
  ggplot(aes(x = month_num, y = value, color = factor(era_start), group = era_start)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  facet_wrap(vars(metric, name), scales = "free_y", ncol = length(plot_areas)) +
  scale_x_continuous(breaks = c(1, 4, 7, 10), labels = month.abb[c(1, 4, 7, 10)]) +
  scale_color_brewer(palette = "Dark2", name = "Year") +
  labs(
    title = "Monthly Water Surface Elevation vs. WET Surface Water — Annual (2022–present)",
    x = "Month", y = NULL
  ) +
  theme_minimal(base_size = 9) +
  theme(legend.position = "bottom", strip.text = element_text(size = 7))
```

## 6.1 Elevation range within each decadal month {#ribbon}

The decadal composites average 10 years of daily elevation into one monthly value, which can
hide how much a reservoir actually swings within that window. Plotting the min–max range
alongside the mean shows that spread directly.

```{r elevation-range-ribbon, fig.height=10}
combined_decadal |>
  filter(name %in% plot_areas) |>
  ggplot(aes(x = month_num, group = era_str)) +
  geom_ribbon(aes(ymin = elevation_min_ft, ymax = elevation_max_ft, fill = era_str), alpha = 0.25) +
  geom_line(aes(y = elevation_mean_ft, color = era_str), linewidth = 1) +
  facet_wrap(~name, scales = "free_y", ncol = 2) +
  scale_x_continuous(breaks = c(1, 4, 7, 10), labels = month.abb[c(1, 4, 7, 10)]) +
  scale_color_brewer(palette = "Set2", name = "Era") +
  scale_fill_brewer(palette = "Set2", name = "Era") +
  labs(
    title    = "Decadal Elevation: Monthly Mean with Min–Max Range",
    subtitle = "Ribbon spans the min/max daily elevation observed within each decadal-month window",
    x = "Month", y = "Water surface elevation (ft)"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r}
knitr::knit_exit()
```

---

# 7. Session info

```{r session-info}
sessionInfo()
```
