---
title: "Refuge Crosswalk: USBR Hydromet/USGS Elevations & WET Wetland Extent"
author: "Maddee Wiggins"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

```{r libraries}
library(tidyverse)
library(sf)
library(ggplot2)
library(scales)
library(patchwork)
library(knitr)
```

# Overview

This document crosswalks water-surface-**elevation** stations from `teacup_diagram_data` (https://www.usbr.gov/pn/hydromet/klamath/teacup.html) to each of the six
Klamath Basin wildlife refuges in (stored in `rivermile::klamath_refuges`), then joins each matched
station's monthly water level to the processed WET (Intermountain West Joint Venture (IWJV)) monthly surface water extent data (`wet_surface_water`) for that refuge. The goal is a combined monthly view — station water
level alongside wetland surface water extent — for refuges with a relevant elevation gauge.

The refuge/station crosswalk is a hand-curated lookup (Section 3), only stations that are actually on or clearly hydrologically tied to a refuge are assigned to it. Three refuges — Bear Valley, Klamath Marsh, and Lower Klamath/Sheepy have no station assigned and are excluded from the combined analysis. 

---

# 1. Load data

```{r load-data}
data("wet_surface_water",    package = "klamathWaterData")
data("teacup_diagram_data",  package = "klamathWaterData")

refuges <- rivermile::klamath_refuges

cat("wet_surface_water:  ", nrow(wet_surface_water),   "rows\n")
cat("teacup_diagram_data:", nrow(teacup_diagram_data), "rows\n")
cat("refuges:            ", nrow(refuges),             "features\n")

kable(st_drop_geometry(refuges), caption = "rivermile::klamath_refuges")
```

---

# 2. Elevation stations in `teacup_diagram_data`

`teacup_diagram_data$measure_type` is one of `elevation`, `storage`, or `flow`. We keep only
`elevation` — reservoir/lake water-surface elevation and gauge height, which is what can
plausibly track a refuge's wetland extent. `lat`/`long` are already included in this dataset
(sourced from the USBR RISE API for USBR Hydromet sites, and `dataRetrieval::readNWISsite()`
for USGS sites — see `data-raw/data-pull/teacup-diagram-data-pull.R`).

```{r elevation-stations}
elev_stations <- teacup_diagram_data |>
  filter(measure_type == "elevation") |>
  distinct(source, site, label, lat, long)

stations_sf <- elev_stations |>
  st_as_sf(coords = c("long", "lat"), crs = 4326, remove = FALSE)

kable(elev_stations, caption = "Elevation stations available in teacup_diagram_data")
```

# 3. Spatial crosswalk of elevation sites with refuges

Each elevation station is assigned to a refuge only where it is genuinely on-site or the
obvious controlling structure — not just the nearest point on a map:

- **Tule Lake NWR** — `TULC` and `TULC2`, the refuge's own Sump 1A and Sump 1B forebay gauges.
- **Clear Lake NWR** — `CLK`, Clear Lake Reservoir West Lobe, inside the refuge.
- **Upper Klamath Lake NWR** — `11507001`, the USGS lake-elevation gauge near Link River Dam.
- **Bear Valley, Klamath Marsh, Lower Klamath/Sheepy** — no station assigned

```{r crosswalk-table}
crosswalk <- tribble(
  ~orgname,                                 ~site,
  "tule lake national wildlife refuge",     "TULC",
  "tule lake national wildlife refuge",     "TULC2",
  "upper klamath national wildlife refuge", "11507001",
  "clear lake national wildlife refuge",    "CLK"
)

stations_refuge_cross <- stations_sf |>
  left_join(crosswalk, by = "site")

stations_refuge_cross |>
  st_drop_geometry() |>
  kable(caption = "Refuge × station crosswalk (NA = elevation station not assigned to a refuge)")
```

## 3.1 Map: refuges & elevation stations

```{r crosswalk-map, fig.height=8}
ggplot() +
  geom_sf(data = refuges, aes(fill = orgname), alpha = 0.5, color = "grey30") +
  geom_sf(data = stations_refuge_cross, color = "black", size = 2) +
  geom_sf_text(data = stations_refuge_cross, aes(label = site), nudge_y = 0.03, size = 3) +
  scale_fill_brewer(palette = "Set3", name = "Refuge") +
  labs(
    title = "Klamath Refuges & Teacup Diagram Elevation Stations",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", legend.text = element_text(size = 7)) +
  guides(fill = guide_legend(nrow = 3))
```

---

# 4. Monthly water level & WET surface water extent

`wet_surface_water` has two temporal structures: decadal composites (`is_decadal == TRUE`, one
value per calendar month averaged over a 10-year era) and individual calendar months from 2022
onward (`is_decadal == FALSE`, `era_start` is the actual year). We match each structure to the
elevation record at the same resolution: a decadal climatological monthly mean for decadal rows,
and the actual year-month mean for annual rows. Where a refuge has multiple stations (Tule
Lake's two sumps), their monthly values are averaged together first. Alongside the mean we also
keep the min/max elevation observed within each aggregation window, used for the range ribbon
in [Section 6.1](#ribbon).

```{r monthly-refuge-xwalk}
refuge_key_lookup <- tribble(
  ~refuge,                ~orgname,
  "upper_klamath_lake",   "upper klamath national wildlife refuge",
  "clear_lake",           "clear lake national wildlife refuge",
  "tule_lake",            "tule lake national wildlife refuge",
  "bear_valley",          "bear valley national wildlife refuge",
  "lower_klamath_sheepy", "lower klamath national wildlife refuge",
  "klamath_marsh",        "klamath marsh national wildlife refuge"
)

elev_xwalk_data <- teacup_diagram_data |>
  filter(measure_type == "elevation") |>
  # many-to-many is expected: TULC and TULC2 both map to tule_lake
  left_join(stations_refuge_cross |> st_drop_geometry() |> select(site, orgname),
            by = "site", relationship = "many-to-many") |>
  left_join(refuge_key_lookup, by = "orgname") |>
  filter(!is.na(refuge))

elev_monthly_decadal <- elev_xwalk_data |>
  mutate(year = year(date), month_num = month(date)) |>
  mutate(era_str = case_when(
    year >= 2005 & year <= 2014 ~ "2005-2014",
    year >= 2015 & year <= 2022 ~ "2015-2022",
    TRUE ~ NA_character_
  )) |>
  filter(!is.na(era_str)) |>
  group_by(refuge, era_str, month_num) |>
  summarise(elevation_mean_ft = mean(value, na.rm = TRUE),
            elevation_min_ft  = min(value, na.rm = TRUE),
            elevation_max_ft  = max(value, na.rm = TRUE), .groups = "drop") |>
  mutate(is_decadal = TRUE)

elev_monthly_annual <- elev_xwalk_data |>
  mutate(year = year(date), month_num = month(date)) |>
  filter(year >= 2022) |>
  group_by(refuge, era_start = year, month_num) |>
  summarise(elevation_mean_ft = mean(value, na.rm = TRUE),
            elevation_min_ft  = min(value, na.rm = TRUE),
            elevation_max_ft  = max(value, na.rm = TRUE), .groups = "drop") |>
  mutate(is_decadal = FALSE)

combined_decadal <- wet_surface_water |>
  filter(is_decadal) |>
  inner_join(elev_monthly_decadal, by = c("refuge", "is_decadal", "era_str", "month_num"))

combined_annual <- wet_surface_water |>
  filter(!is_decadal) |>
  inner_join(elev_monthly_annual, by = c("refuge", "is_decadal", "era_start", "month_num"))

combined_monthly <- bind_rows(combined_decadal, combined_annual) |>
  arrange(refuge, is_decadal, era_start, month_num)

plot_refuges <- sort(unique(combined_monthly$refuge))

cat("Combined monthly rows (decadal):", nrow(combined_decadal), "\n")
cat("Combined monthly rows (annual): ", nrow(combined_annual), "\n")
cat("Refuges represented:", paste(plot_refuges, collapse = ", "), "\n")

kable(head(combined_monthly, 10), caption = "combined_monthly — first rows")
```

## 4.1 Package dataset: `refuge_elevation_wetland`

`combined_decadal` and `combined_annual` are working objects scoped to this document. For
downstream use in the package, they're combined (via `combined_monthly`) into a single, tidy,
renamed dataset — one row per refuge-month, elevation and WET columns clearly distinguished by
prefix — and saved as `refuge_elevation_wetland`.

```{r package-dataset}
refuge_elevation_wetland <- combined_monthly |>
  transmute(
    refuge,
    is_decadal,
    era_str,
    era_start,
    month_num,
    month_chr,
    elevation_mean_ft,
    elevation_min_ft,
    elevation_max_ft,
    wet_mean_val   = mean_val,
    wet_median_val = median_val,
    wet_sd_val     = sd_val,
    wet_n_pixels   = n_pixels
  ) |>
  arrange(refuge, is_decadal, era_start, month_num) |> 
  mutate(month_chr = tolower(month_chr)) |>
  select(-era_start)

glimpse(refuge_elevation_wetland)
kable(head(refuge_elevation_wetland, 10), caption = "refuge_elevation_wetland — first rows")

usethis::use_data(refuge_elevation_wetland, overwrite = TRUE)
```

---

# 5. Combined monthly plots

Dual view of station elevation and WET mean surface water value for the three refuges with an
assigned station: Clear Lake, Tule Lake, and Upper Klamath Lake.

```{r combined-decadal-plot, fig.height=8}
combined_decadal |>
  filter(refuge %in% plot_refuges) |>
  select(refuge, era_str, month_num, elevation_mean_ft, mean_val) |>
  pivot_longer(c(elevation_mean_ft, mean_val), names_to = "metric", values_to = "value") |>
  mutate(metric = recode(metric,
                         elevation_mean_ft = "Water surface elevation (ft)",
                         mean_val           = "WET mean surface water value")) |>
  ggplot(aes(x = month_num, y = value, color = era_str, group = era_str)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  facet_wrap(vars(metric, refuge), scales = "free_y", ncol = length(plot_refuges)) +
  scale_x_continuous(breaks = c(1, 4, 7, 10), labels = month.abb[c(1, 4, 7, 10)]) +
  scale_color_brewer(palette = "Set2", name = "Era") +
  labs(
    title = "Monthly Water Surface Elevation vs. WET Surface Water — Decadal Composites",
    x = "Month", y = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", strip.text = element_text(size = 8))
```

Annual monthly values (2022–present):

```{r combined-annual-plot, fig.height=8}
combined_annual |>
  filter(refuge %in% plot_refuges) |>
  select(refuge, era_start, month_num, elevation_mean_ft, mean_val) |>
  pivot_longer(c(elevation_mean_ft, mean_val), names_to = "metric", values_to = "value") |>
  mutate(metric = recode(metric,
                         elevation_mean_ft = "Mean water surface elevation (ft)",
                         mean_val           = "WET mean surface water value")) |>
  ggplot(aes(x = month_num, y = value, color = factor(era_start), group = era_start)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  facet_wrap(vars(metric, refuge), scales = "free_y", ncol = length(plot_refuges)) +
  scale_x_continuous(breaks = c(1, 4, 7, 10), labels = month.abb[c(1, 4, 7, 10)]) +
  scale_color_brewer(palette = "Dark2", name = "Year") +
  labs(
    title = "Monthly Water Surface Elevation vs. WET Surface Water — Annual (2022–present)",
    x = "Month", y = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", strip.text = element_text(size = 8))
```

## 5.1 Elevation range within each decadal month {#ribbon}

The decadal composites average 10 years of daily elevation into one monthly value, which can
hide how much a reservoir actually swings within that window. Plotting the min–max range
alongside the mean shows that spread directly.

```{r elevation-range-ribbon, fig.height=7}
combined_decadal |>
  filter(refuge %in% plot_refuges) |>
  ggplot(aes(x = month_num, group = era_str)) +
  geom_ribbon(aes(ymin = elevation_min_ft, ymax = elevation_max_ft, fill = era_str), alpha = 0.25) +
  geom_line(aes(y = elevation_mean_ft, color = era_str), linewidth = 1) +
  facet_wrap(~refuge, scales = "free_y", ncol = 1) +
  scale_x_continuous(breaks = c(1, 4, 7, 10), labels = month.abb[c(1, 4, 7, 10)]) +
  scale_color_brewer(palette = "Set2", name = "Era") +
  scale_fill_brewer(palette = "Set2", name = "Era") +
  labs(
    title    = "Decadal Elevation: Monthly Mean with Min–Max Range",
    subtitle = "Ribbon spans the min/max daily elevation observed within each decadal-month window",
    x = "Month", y = "Water surface elevation (ft)"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r}
knitr::knit_exit()
```

---

# 6. Session info

```{r session-info}
sessionInfo()
```
