---
title: "Refuge Crosswalk: USBR Hydromet Elevations & WET Wetland Extent"
author: "Maddee Wiggins"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

```{r libraries}
library(tidyverse)
library(sf)
library(ggplot2)
library(scales)
library(patchwork)
library(knitr)
```

# Overview

This document crosswalks USBR Hydromet **elevation** (and gauge-height/stage) stations to
each of the six Klamath Basin wildlife refuges in `rivermile::klamath_refuges`, then joins
the nearest hydrologically relevant station's monthly water level to the processed WET
**monthly surface water extent** data (`wet_surface_water`) for that refuge. The goal is a
combined monthly view — station water level alongside wetland surface water extent — for
refuges where a nearby Hydromet gauge exists.

The crosswalk is **proximity-based**: for each refuge we compute the distance from every
Hydromet elevation station to the refuge boundary and rank candidates. Proximity is a
reasonable first-pass heuristic but is not a guarantee of hydrologic connection — see
[Section 6](#gaps) for refuges where the nearest station is too far away (or not actually
the controlling structure) to be considered relevant.

---

# 1. Load data

```{r load-data}
data("wet_surface_water", package = "klamathWaterData")
data("hydromet_data",     package = "klamathWaterData")

refuges <- rivermile::klamath_refuges

cat("wet_surface_water:", nrow(wet_surface_water), "rows\n")
cat("hydromet_data:    ", nrow(hydromet_data),     "rows\n")
cat("refuges:          ", nrow(refuges),           "features\n")

kable(st_drop_geometry(refuges), caption = "rivermile::klamath_refuges")
```

---

# 2. USBR Hydromet elevation stations

`hydromet_data$parameter` mixes elevation/stage parameters with flow parameters. We keep
only the elevation- and gauge-height-type parameters — flow (`Discharge...`,
`Canal/diversion discharge`, `Pumped discharge`) is out of scope for this crosswalk.

```{r elevation-params}
elevation_params <- c(
  "Forebay elevation (daily average)",
  "Forebay elevation (instantaneous)",
  "Gauge height (daily average)",
  "Gauge height, junction/sump",
  "Gauge height, junction/sump #2"
)

hydromet_elev <- hydromet_data |> filter(parameter %in% elevation_params)

kable(
  hydromet_elev |> distinct(site, parameter, label, units),
  caption = "USBR Hydromet elevation/stage parameters available"
)

no_elev <- setdiff(unique(hydromet_data$site), unique(hydromet_elev$site))
cat("Sites with NO elevation-type parameter (flow-only, excluded):",
    paste(no_elev, collapse = ", "), "\n")
```

## 2.1 Station coordinates

`hydromet_data` does not carry station coordinates, so this lookup was built from the USBR
RISE open-data API (`data.usbr.gov/rise/api/location`), which catalogs each Hydromet `cbtt`
under a named Reclamation location.

```{r station-lookup}
elev_station_lookup <- tribble(
  ~site,  ~lat,      ~long,       ~station_name,
  "LRS",  41.926030, -121.075876, "Clear Lake Reservoir East Lobe",
  "CLK",  41.846626, -121.209623, "Clear Lake Reservoir West Lobe",
  "LRSC", 41.927834, -121.074174, "Lost River below Clear Lake Dam",
  "KELO", 42.163299, -121.319785, "Lost River near E Langell Valley Rd Bridge",
  "HRPO", 42.169507, -121.453355, "Lost River at Harpold Dam",
  "GER",  42.201271, -121.129795, "Gerber Reservoir and Dam",
  "MAL",  42.005769, -121.223952, "Malone Diversion Reservoir and Dam",
  "EEEO", 42.047187, -121.795794, "Klamath Straits Drain Pumping Plant E/EE",
  "FPPO", 42.079884, -121.843742, "Klamath Straits Drain Pumping Plant F/FF",
  "STDC", 41.996724, -121.777916, "Klamath Straits Drain at Headworks (Hwy 161)"
)

stations_sf <- elev_station_lookup |>
  st_as_sf(coords = c("long", "lat"), crs = 4326, remove = FALSE)

kable(elev_station_lookup, caption = "USBR Hydromet elevation station coordinates")
```

---

# 3. Spatial crosswalk: nearest station per refuge

Distances are computed in a projected CRS (EPSG:32610, UTM Zone 10N — meters, appropriate
for the Klamath Basin) from each station to the refuge polygon boundary, then converted to
miles.

```{r crosswalk-distance}
BUFFER_CRS <- 32610

stations_proj <- st_transform(stations_sf, BUFFER_CRS)
refuges_proj  <- st_transform(refuges, BUFFER_CRS)

dist_m <- st_distance(stations_proj, refuges_proj)

crosswalk_dist <- as_tibble(dist_m) |>
  setNames(refuges_proj$orgname) |>
  mutate(site = stations_proj$site) |>
  pivot_longer(-site, names_to = "orgname", values_to = "dist_m") |>
  mutate(dist_mi = as.numeric(dist_m) / 1609.344) |>
  left_join(elev_station_lookup |> select(site, station_name), by = "site") |>
  arrange(orgname, dist_mi)

nearest_per_refuge <- crosswalk_dist |>
  group_by(orgname) |>
  slice_min(dist_mi, n = 3) |>
  ungroup()

kable(
  nearest_per_refuge |> mutate(dist_mi = round(dist_mi, 2)) |>
    select(orgname, site, station_name, dist_mi),
  caption = "Three nearest USBR Hydromet elevation stations per refuge"
)
```

## 3.1 Map: refuges & elevation stations

```{r crosswalk-map, fig.height=8}
ggplot() +
  geom_sf(data = refuges, aes(fill = orgname), alpha = 0.5, color = "grey30") +
  geom_sf(data = stations_sf, color = "black", size = 2) +
  geom_sf_text(data = stations_sf, aes(label = site), nudge_y = 0.02, size = 3) +
  scale_fill_brewer(palette = "Set3", name = "Refuge") +
  labs(
    title = "Klamath Refuges & USBR Hydromet Elevation Stations",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", legend.text = element_text(size = 7)) +
  guides(fill = guide_legend(nrow = 3))
```

---

# 4. Selecting a representative station per refuge

For refuges with a station reasonably close to the boundary, we take the single nearest
station as the "hydrologically relevant" gauge. Where a site has more than one elevation-type
parameter (e.g. `GER` has both forebay elevation and gauge height), we prefer, in order:
forebay elevation (daily average) > gauge height (daily average) > gauge height, junction/sump
> forebay elevation (instantaneous) > gauge height, junction/sump #2.

```{r select-station}
param_priority <- c(
  "Forebay elevation (daily average)",
  "Gauge height (daily average)",
  "Gauge height, junction/sump",
  "Forebay elevation (instantaneous)",
  "Gauge height, junction/sump #2"
)

refuge_station_xwalk <- nearest_per_refuge |>
  group_by(orgname) |>
  slice_min(dist_mi, n = 1) |>
  ungroup() |>
  select(orgname, site, station_name, dist_mi)

refuge_param_xwalk <- refuge_station_xwalk |>
  left_join(hydromet_elev |> distinct(site, parameter), by = "site") |>
  mutate(param_rank = match(parameter, param_priority)) |>
  group_by(orgname) |>
  slice_min(param_rank, n = 1) |>
  ungroup() |>
  select(orgname, site, station_name, parameter, dist_mi)

kable(
  refuge_param_xwalk |> mutate(dist_mi = round(dist_mi, 2)),
  caption = "Selected USBR Hydromet elevation series per refuge"
)
```

---

# 5. Monthly water level & WET surface water extent

`wet_surface_water` has two temporal structures: decadal composites (`is_decadal == TRUE`,
one value per calendar month averaged over a 10-year era) and individual calendar months
from 2022 onward (`is_decadal == FALSE`, `era_start` is the actual year). We match each
structure to the Hydromet record at the same resolution: a decadal climatological monthly
mean for decadal rows, and the actual year-month mean for annual rows.

```{r monthly-refuge-xwalk}
refuge_key_lookup <- tribble(
  ~refuge,                ~orgname,
  "upper_klamath_lake",   "upper klamath national wildlife refuge",
  "clear_lake",           "clear lake national wildlife refuge",
  "tule_lake",            "tule lake national wildlife refuge",
  "bear_valley",          "bear valley national wildlife refuge",
  "lower_klamath_sheepy", "lower klamath national wildlife refuge",
  "klamath_marsh",        "klamath marsh national wildlife refuge"
)

xwalk <- refuge_param_xwalk |>
  left_join(refuge_key_lookup, by = "orgname")

hydromet_monthly_decadal <- hydromet_elev |>
  inner_join(xwalk |> select(refuge, site, parameter), by = c("site", "parameter")) |>
  mutate(year = year(date), month_num = month(date)) |>
  # match wet_surface_water's decadal eras
  mutate(era_str = case_when(
    year >= 2005 & year <= 2014 ~ "2005-2014",
    year >= 2015 & year <= 2022 ~ "2015-2022",
    TRUE ~ NA_character_
  )) |>
  filter(!is.na(era_str)) |>
  group_by(refuge, site, era_str, month_num) |>
  summarise(elevation_ft = mean(value, na.rm = TRUE), .groups = "drop") |>
  mutate(is_decadal = TRUE)

hydromet_monthly_annual <- hydromet_elev |>
  inner_join(xwalk |> select(refuge, site, parameter), by = c("site", "parameter")) |>
  mutate(year = year(date), month_num = month(date)) |>
  filter(year >= 2022) |>
  group_by(refuge, site, era_start = year, month_num) |>
  summarise(elevation_ft = mean(value, na.rm = TRUE), .groups = "drop") |>
  mutate(is_decadal = FALSE)

combined_decadal <- wet_surface_water |>
  filter(is_decadal) |>
  inner_join(hydromet_monthly_decadal, by = c("refuge", "is_decadal", "era_str", "month_num"))

combined_annual <- wet_surface_water |>
  filter(!is_decadal) |>
  inner_join(hydromet_monthly_annual, by = c("refuge", "is_decadal", "era_start", "month_num"))

combined_monthly <- bind_rows(combined_decadal, combined_annual) |>
  arrange(refuge, is_decadal, era_start, month_num)

cat("Combined monthly rows (decadal):", nrow(combined_decadal), "\n")
cat("Combined monthly rows (annual): ", nrow(combined_annual), "\n")
cat("Refuges represented:", paste(unique(combined_monthly$refuge), collapse = ", "), "\n")

kable(head(combined_monthly, 10), caption = "combined_monthly — first rows")
```

---

# 6. Data gaps & caveats {#gaps}

Not every refuge has a Hydromet elevation station close enough to be considered
hydrologically relevant. Distances to the nearest candidate:

```{r gap-summary}
gap_summary <- refuge_station_xwalk |>
  mutate(
    relevance = case_when(
      dist_mi < 2  ~ "Direct / on-site",
      dist_mi < 12 ~ "Nearby, plausible",
      TRUE         ~ "Too far — not relevant"
    )
  )

kable(
  gap_summary |> mutate(dist_mi = round(dist_mi, 2)),
  caption = "Nearest station relevance by refuge"
)
```

- **Clear Lake** and **Lower Klamath/Sheepy** have a station essentially on-site (`CLK` sits
  inside Clear Lake Reservoir; `STDC` sits inside the Lower Klamath refuge boundary at the
  Straits Drain headworks) — the crosswalk is direct and reliable here.
- **Tule Lake** and **Bear Valley** have their nearest station several miles away, and in both
  cases the nearest station is Klamath Straits Drain infrastructure (`STDC`/`EEEO`/`FPPO`),
  which serves the Tule Lake/Lower Klamath conveyance system rather than being *inside* either
  refuge. Treat the Tule Lake and Bear Valley rows in `combined_monthly` as a loose regional
  proxy, not a direct on-site water level. Tule Lake's actual controlling structures are the
  Tule Lake Sump 1A/1B forebay gauges (`TULC`/`TULC2`), which are not present in this version
  of `hydromet_data` (see `data-raw/data-pull/teacup-diagram-data-pull.R` on a separate branch
  for a pull that includes them).
- **Klamath Marsh** has no Hydromet elevation station within a reasonable distance (nearest is
  ~46 miles). It is excluded from `combined_monthly` entirely — there is no relevant USBR
  Hydromet gauge for this refuge in the current data.
- **Upper Klamath Lake** is also excluded here for the same reason (nearest Hydromet elevation
  station is ~15 miles away, at Klamath Straits Drain infrastructure that is not on the lake
  itself). Upper Klamath Lake's real elevation record is a USGS product already in this
  package as `water_level_data`/`water_level_gage` (gages 11505900, 11504300, 11505800,
  11507001), not USBR Hydromet — a natural follow-up would repeat this crosswalk against that
  dataset for Upper Klamath Lake specifically.
- Recall from `wet-process-data.R` / `R/data.R` that **Bear Valley** also has no WET-detected
  surface water within its own refuge boundary in any product or era (visual QC confirmed
  this separately) — so even where a proxy elevation series exists for Bear Valley, there is
  no wetland extent signal on the other side of the join for it to move with.

---

# 7. Combined monthly plots

Refuges with a direct or plausible station only (`Clear Lake`, `Lower Klamath/Sheepy`,
`Tule Lake`) — dual view of station elevation/stage and WET mean surface water value, decadal
composites.

```{r combined-decadal-plot, fig.height=8}
plot_refuges <- c("clear_lake", "lower_klamath_sheepy", "tule_lake")

combined_decadal |>
  filter(refuge %in% plot_refuges) |>
  select(refuge, era_str, month_num, elevation_ft, mean_val) |>
  pivot_longer(c(elevation_ft, mean_val), names_to = "metric", values_to = "value") |>
  mutate(metric = recode(metric,
                         elevation_ft = "Hydromet elevation/stage (ft)",
                         mean_val     = "WET mean surface water value")) |>
  ggplot(aes(x = month_num, y = value, color = era_str, group = era_str)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  facet_grid(metric ~ refuge, scales = "free_y") +
  scale_x_continuous(breaks = c(1, 4, 7, 10), labels = month.abb[c(1, 4, 7, 10)]) +
  scale_color_brewer(palette = "Set2", name = "Era") +
  labs(
    title = "Monthly Hydromet Elevation vs. WET Surface Water — Decadal Composites",
    x = "Month", y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

Annual monthly values (2022–present):

```{r combined-annual-plot, fig.height=8}
combined_annual |>
  filter(refuge %in% plot_refuges) |>
  select(refuge, era_start, month_num, elevation_ft, mean_val) |>
  pivot_longer(c(elevation_ft, mean_val), names_to = "metric", values_to = "value") |>
  mutate(metric = recode(metric,
                         elevation_ft = "Hydromet elevation/stage (ft)",
                         mean_val     = "WET mean surface water value")) |>
  ggplot(aes(x = month_num, y = value, color = factor(era_start), group = era_start)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  facet_grid(metric ~ refuge, scales = "free_y") +
  scale_x_continuous(breaks = c(1, 4, 7, 10), labels = month.abb[c(1, 4, 7, 10)]) +
  scale_color_brewer(palette = "Dark2", name = "Year") +
  labs(
    title = "Monthly Hydromet Elevation vs. WET Surface Water — Annual (2022–present)",
    x = "Month", y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

## 7.1 Elevation vs. surface water extent — scatter

```{r elevation-vs-wet-scatter, fig.height=6}
combined_monthly |>
  filter(refuge %in% plot_refuges) |>
  ggplot(aes(x = elevation_ft, y = mean_val, color = refuge)) +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  facet_wrap(~refuge, scales = "free") +
  scale_color_brewer(palette = "Set2", guide = "none") +
  labs(
    title = "Hydromet Elevation/Stage vs. WET Mean Surface Water Value",
    subtitle = "One point per refuge-month (decadal + annual)",
    x = "Hydromet elevation/stage (ft)",
    y = "WET mean surface water value"
  ) +
  theme_minimal(base_size = 12)
```

```{r elevation-vs-wet-cor}
combined_monthly |>
  filter(refuge %in% plot_refuges) |>
  group_by(refuge) |>
  summarise(
    n = n(),
    cor_elevation_meanval = cor(elevation_ft, mean_val, use = "complete.obs"),
    cor_elevation_pctwet  = cor(elevation_ft, pct_wet,  use = "complete.obs"),
    .groups = "drop"
  ) |>
  kable(caption = "Correlation between Hydromet elevation/stage and WET metrics, by refuge")
```

---

# 8. Session info

```{r session-info}
sessionInfo()
```
