---
title: "Refuges Water Quality Data Exploration (Lower Klamath and Tule Lake NWR)"
author: "Badhia Yunes Katz"
date: "October 2026"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 9, fig.height = 4.5)
library(tidyverse)
library(readxl)
library(janitor)
library(lubridate)
library(leaflet)
library(plotly)
library(htmltools)
library(knitr)
```

```{r read}
f <- "data-raw/skinner-wq-data/Refuges water quality data for Recovery.xlsx"
excel_date <- function(x) as.Date(as.numeric(x), origin = "1899-12-30")
fmt <- function(d) format(d, "%d %b %Y")

# "NA" is stored as literal text in many cells, so everything is read as text first
read_txt <- function(sheet) {
  read_excel(f, sheet = sheet, col_types = "text", .name_repair = "minimal") |>
    clean_names() |>
    mutate(across(everything(), ~ na_if(.x, "NA")))
}
fix_site <- function(x) str_replace(x, "-", "_")   # site ids mix "-" and "_" (e.g. TLNWR1a-NL vs TLNWR1a_NL)

sonde_cols <- c("temp", "p_h", "do_conc", "do_sat", "sp_cond", "tds", "sal", "depth", "bat", "cab")
nut_cols   <- c("tdp", "tp", "po4", "nh3", "n_ox", "tn", "pp", "per_pp", "n_p")
tox_cols   <- c("atx", "cyl", "mc", "stx")

ysi <- read_txt("YSI") |>
  mutate(date = excel_date(date), site = fix_site(site), across(all_of(sonde_cols), as.numeric))
cont <- ysi |> filter(type == "Continuous")

nc <- read_txt("Nutrient and cyano data ") |>
  mutate(date = excel_date(date), site = fix_site(site), across(all_of(c(nut_cols, tox_cols)), as.numeric))

# site coordinates come from the two site tabs and from the coordinates recorded in the data
sites_2024 <- read_excel(f, sheet = "2024+ sampling sites", .name_repair = "minimal") |>
  clean_names() |> transmute(site = fix_site(site_name), description = site_description, coordinates)
sites_2021 <- read_excel(f, sheet = "2021-2023 cyano sampling sites", .name_repair = "minimal") |>
  clean_names() |> filter(!is.na(coordinates)) |>
  transmute(site = fix_site(site_id), description = NA_character_, coordinates)
sites_data <- nc |> filter(!is.na(coordinates)) |>
  distinct(site, .keep_all = TRUE) |> select(site, description = site_description, coordinates)

site_coords <- bind_rows(sites_2024, sites_2021, sites_data) |>
  distinct(site, .keep_all = TRUE) |>
  separate(coordinates, c("latitude", "longitude"), sep = ",", convert = TRUE) |>
  mutate(latitude = as.numeric(latitude), longitude = as.numeric(longitude))

site_tbl <- bind_rows(ysi |> transmute(site, date, dataset = paste0("sonde (", str_to_lower(type), ")")),
                      nc  |> transmute(site, date, dataset = "nutrient/cyano")) |>
  group_by(site) |>
  summarise(sonde_days = n_distinct(date[dataset == "sonde (continuous)"]),
            discrete_readings = sum(dataset == "sonde (discrete)"),
            nutrient_cyano_samples = sum(dataset == "nutrient/cyano"),
            first = min(date), last = max(date), .groups = "drop") |>
  left_join(site_coords, by = "site") |>
  mutate(refuge = if_else(str_starts(site, "LKNWR"), "Lower Klamath", "Tule Lake")) |>
  arrange(refuge, site)

cont_sites <- site_tbl$site[site_tbl$sonde_days > 0]

# one row per site / year of continuous sonde data
deployments <- cont |>
  group_by(site, year) |>
  summarise(first = min(date), last = max(date), days = n_distinct(date), .groups = "drop") |>
  arrange(site, first)

neg_do <- round(100 * mean(cont$do_conc < 0, na.rm = TRUE))
```

This report describes the "Refuges water quality data for Recovery" workbook
(`data-raw/skinner-wq-data/Refuges water quality data for Recovery.xlsx`) from Lower Klamath and Tule Lake National Wildlife Refuges:
where the data were collected, what was measured, over what period, and what to watch out for when using it.

# At a glance

| | |
|---|---|
| **Where** | `r nrow(site_tbl)` sites on two refuges: `r sum(site_tbl$refuge == "Lower Klamath")` at Lower Klamath (Sheepy Lake area) and `r sum(site_tbl$refuge == "Tule Lake")` at Tule Lake (sumps 1A and 1B) |
| **Continuous sonde data** | Hourly readings at `r length(cont_sites)` sites, `r fmt(min(cont$date))` to `r fmt(max(cont$date))`, in seasonal deployments |
| **Nutrient and cyanotoxin samples** | `r nrow(nc)` samples, `r fmt(min(nc$date))` to `r fmt(max(nc$date))`, at `r n_distinct(nc$site)` sites (nutrients at one site only) |
| **Context series** | Daily flow (Anderson Rose and Ady Canal) and Unit 2 water surface elevation |
| **Main gaps** | No continuous data in winter, none in 2021, 2022 or 2026; sampling dropped sharply after April 2023 |

# 1. Where the data were collected

```{r cont-sites}
site_tbl |>
  filter(sonde_days > 0) |>
  left_join(deployments |> group_by(site) |> summarise(Period = paste(fmt(min(first)), "-", fmt(max(last)))), by = "site") |>
  transmute(Site = site, Refuge = refuge, Description = coalesce(description, "(no description given)"),
            Latitude = latitude, Longitude = longitude, `Continuous sonde period` = Period) |>
  kable(caption = "Sites with continuous sonde data")
```

```{r map}
site_tbl |>
  drop_na(latitude, longitude) |>
  mutate(has_sonde = sonde_days > 0) |>
  leaflet() |>
  addProviderTiles("Esri.WorldImagery") |>
  addCircleMarkers(~longitude, ~latitude, radius = 7, color = "white", weight = 1,
                   fillColor = ~if_else(has_sonde, "orange", "steelblue"), fillOpacity = 0.9,
                   label = ~site,
                   popup = ~paste0("<b>", site, "</b><br>", coalesce(description, ""),
                                   "<br>Continuous sonde days: ", sonde_days,
                                   "<br>Nutrient/cyano samples: ", nutrient_cyano_samples)) |>
  addLegend("bottomright", colors = c("orange", "steelblue"),
            labels = c("Continuous sonde site", "Discrete sampling only"), opacity = 1)
```

- **Routine cyanotoxin sites (2021-2023):** Sheepy Lake (`LKNWR_SLBR`, `LKNWR_WSL`) and Tule Lake sumps 1A/1B (`TLNWR1a_BR`, `_NL`, `_SW`, `TLNWR1b_NW`). After April 2023 only a few sites were sampled, mostly during blooms; `TLNWR1a_DH` ("Donut Hole") continues to 2026.
- **One-off sites:** `LKNWR_U3`, `TLNWR1a_EN`, `_SPR`, `_STN`, bottom-of-water-column samples (`_bot`) and a few early `TLNWR1a_1/2/3` and `_EC` samples.
- **Missing coordinates:** `r paste(site_tbl$site[is.na(site_tbl$latitude)], collapse = ", ")` (not shown on the map).

# 2. What was measured

```{r param-table}
n_nut <- sum(!is.na(nc$tp))
tibble(
  Data = c("Sonde (hourly)", "Nutrients (lab)", "Cyanotoxins", "Cyanobacteria taxa", "Context (daily)"),
  Parameters = c("Temperature, pH, dissolved oxygen (concentration and saturation), specific conductance, TDS, salinity, depth",
                 "Total dissolved P, total P, orthophosphate, ammonia, nitrate/nitrite, total N",
                 "Microcystin, saxitoxin, anatoxin-a, cylindrospermopsin",
                 "Qualitative description of taxa present in the toxin sample",
                 "Anderson Rose flow, Ady Canal flow, Unit 2 water surface elevation"),
  Units = c("°C, std units, mg/L, %, uS/cm, mg/L, psu, m (depth is relative, not accurate)", "ug/L", "ug/L", "free text",
            "flow units not stated; elevation in ft"),
  Coverage = c(paste(length(cont_sites), "sites, hourly"),
               paste(n_nut, "samples, all at TLNWR1a_DH"),
               paste(sum(!is.na(nc$mc)), "samples at", n_distinct(nc$site[!is.na(nc$mc)]), "sites"),
               "most cyano samples", "see section 6")
) |> kable()
```

A `0` in a nutrient or toxin column means **not detected / below the detection limit**, not a true zero. Per the workbook, nutrient analyses
were done by the SRWQL lab. Particulate P, percent particulate P and the N:P ratio are calculated from the lab results.

# 3. Time period and gaps

```{r deployment-bullets, results = "asis"}
# one line per continuous sonde site listing its seasonal records
deployments |>
  mutate(period = paste0(format(first, "%d %b"), " - ", format(last, "%d %b %Y"))) |>
  group_by(site) |>
  summarise(txt = paste0("- **", first(site), ":** ", paste(period, collapse = "; "))) |>
  pull(txt) |>
  writeLines()
```

**Gaps**

- **No winter data:** continuous records run only from about May to November. `LKNWR_AI` and `TLNWR1a_LRI` have no data from late Nov 2024 to 27 May 2025.
- **No continuous data** in 2021, 2022 or 2026.
- **`TLNWR1a_DH`:** the sonde ended 14 Aug 2024; after that only discrete readings exist (4-6 per year).
- **Nutrient and cyanotoxin samples:** dense in 2021-2022 (`r sum(year(nc$date) == 2021)` and `r sum(year(nc$date) == 2022)` samples), `r sum(year(nc$date) == 2023)` in 2023 (routine sites stopped in April 2023), then only 5-10 a year.
- **Nutrients** were only measured at `TLNWR1a_DH` (`r n_nut` samples, 2024-2026).

# 4. Time series

Each tab shows one continuous sonde site. For readability the hourly readings are summarised to a **daily mean (line) with the daily minimum-maximum range (shaded)**.
Breaks in the line are periods with no data. The plots are interactive: hover for the date and values, drag to zoom (all panels in a tab zoom together) and double-click to reset.
Each panel has its own y-axis.

```{r plotly-deps, echo = FALSE}
# widgets printed inside the results = "asis" loop below do not register their JS dependencies on their own;
# an empty widget here makes knitr include plotly.js in the page
dummy <- plot_ly(x = 1, y = 1, type = "scatter", mode = "markers", marker = list(opacity = 0)) |>
  layout(margin = list(t = 0, b = 0, l = 0, r = 0), xaxis = list(visible = FALSE), yaxis = list(visible = FALSE))
dummy$height <- 1
htmltools::tagList(dummy)
```

```{=html}
<script>
// plotly widgets drawn inside hidden tabs have no width; resize them when their tab is opened
document.addEventListener("click", function (e) {
  if (!e.target.closest(".tabset")) return;
  setTimeout(function () {
    document.querySelectorAll(".plotly.html-widget").forEach(function (w) {
      if (w.offsetParent !== null && window.Plotly) Plotly.Plots.resize(w);
    });
  }, 200);
});
</script>
```

## Time series by site {.tabset .tabset-pills}

```{r tabs, results = "asis"}
par_labels <- c(temp = "Temperature (°C)", p_h = "pH", do_conc = "DO (mg/L)",
                do_sat = "DO saturation (%)", sp_cond = "Sp. conductance (uS/cm)")
par_units  <- c(temp = "°C", p_h = "", do_conc = "mg/L", do_sat = "%", sp_cond = "uS/cm")

# daily mean / min / max; missing calendar days are added as NA so the line breaks at gaps
na_if_inf <- function(x) if_else(is.finite(x), x, NA_real_)
daily <- cont |>
  pivot_longer(all_of(names(par_labels)), names_to = "parameter", values_to = "value") |>
  drop_na(value) |>
  group_by(site, parameter, date) |>
  summarise(mean = mean(value), lo = min(value), hi = max(value), .groups = "drop") |>
  group_by(site, parameter) |>
  complete(date = seq(min(date), max(date), by = "day")) |>
  ungroup()

panel_title <- function(label) {
  list(list(text = paste0("<b>", label, "</b>"), x = 0, y = 1.02, xref = "paper", yref = "paper",
            xanchor = "left", yanchor = "bottom", showarrow = FALSE, font = list(size = 12)))
}

site_panels <- function(s) {
  panels <- imap(par_labels, function(label, par) {
    d <- daily |> filter(site == s, parameter == par) |>
      mutate(txt = sprintf("%s<br><b>%.3g</b> %s (daily mean)<br>range %.3g to %.3g", format(date, "%d %b %Y"), mean, par_units[[par]], lo, hi))
    # runs of consecutive days with data; the min-max band is drawn per run so it does not bridge gaps
    d <- d |> mutate(run = cumsum(!is.na(mean) & is.na(lag(mean, default = NA_real_))))
    p <- plot_ly(showlegend = FALSE)
    for (r in unique(d$run[!is.na(d$mean)])) {
      dr <- filter(d, run == r, !is.na(mean))
      p <- p |>
        add_trace(data = dr, x = ~date, y = ~hi, type = "scatter", mode = "lines", line = list(width = 0, color = "darkorange"), hoverinfo = "skip") |>
        add_trace(data = dr, x = ~date, y = ~lo, type = "scatter", mode = "lines", line = list(width = 0, color = "darkorange"),
                  fill = "tonexty", fillcolor = "rgba(255,140,0,0.25)", hoverinfo = "skip")
    }
    p |>
      add_trace(data = d, x = ~date, y = ~mean, type = "scatter", mode = "lines+markers", text = ~txt, hoverinfo = "text",
                line = list(color = "darkorange3", width = 1.2), marker = list(color = "darkorange3", size = 3)) |>
      layout(annotations = panel_title(label), xaxis = list(title = "", tickfont = list(size = 9), nticks = 8, tickformat = "%b %Y"),
             yaxis = list(title = "", tickfont = list(size = 9)))
  })
  sp <- subplot(unname(panels), nrows = length(panels), margin = c(0.04, 0.04, 0.05, 0.05), shareX = FALSE, titleY = FALSE)
  ax <- paste0("xaxis", c("", 2:length(panels)))
  sp <- do.call(layout, c(list(sp), setNames(rep(list(list(matches = "x")), length(ax) - 1), ax[-1])))
  sp <- sp |> layout(margin = list(t = 30)) |> config(displaylogo = FALSE)
  sp$height <- 950
  sp
}

for (s in cont_sites) {
  desc <- coalesce(site_tbl$description[site_tbl$site == s], "no description given")
  per  <- deployments |> filter(site == s)
  writeLines(c("", "", paste0("### ", s), "", paste0(desc, ". Continuous sonde: ", fmt(min(per$first)), " to ", fmt(max(per$last)), "."), ""))
  print(htmltools::tagList(site_panels(s)))
}
```

## What the plots show {.unlisted .unnumbered}

- **Seasonal pattern:** temperature rises through summer at every site, with the sondes only deployed from spring to autumn.
- **Large daily swings:** the shaded min-max bands show that DO and pH move strongly within a day.
- **Low oxygen:** at `LKNWR_AI` and `TLNWR1a_LRI` DO is often near or below zero in summer and autumn, which is partly a sensor issue (see section 7).

# 5. Nutrient and cyanotoxin samples

- `r nrow(nc)` samples from `r fmt(min(nc$date))` to `r fmt(max(nc$date))`: routine sampling at 6 sites in 2021-2022 (plus a few early one-off sites), then a few sites during blooms.
- Cyanotoxins were measured at most events (`r sum(!is.na(nc$mc))` samples). Microcystin was detected in `r sum(nc$mc > 0, na.rm = TRUE)` of them, with the highest values at `LKNWR_SLBR` (11.7 ug/L), `TLNWR1a_DH` (8.84) and `TLNWR1a_BR` (8.26).
- Nutrients were only measured at `TLNWR1a_DH` (`r n_nut` samples): total P ranged from `r format(min(nc$tp, na.rm = TRUE), big.mark = ",")` to `r format(max(nc$tp, na.rm = TRUE), big.mark = ",")` ug/L and total N from `r format(min(nc$tn, na.rm = TRUE), big.mark = ",")` to `r format(max(nc$tn, na.rm = TRUE), big.mark = ",")` ug/L.
- Observed cyanobacteria genera and bloom notes are in the free-text `taxa` and `notes` columns.

Each tab below is a time series of the individual samples (one point per sample, no averaging). Hover for the site, date and value, drag to zoom
(all panels in a tab zoom together). **Open markers at zero are non-detects** (below the detection limit), not true zeros. The cyanotoxin tab includes all sites; click a site in the legend to hide or show it.

## Samples over time {.tabset .tabset-pills}

```{r nc-tabs, results = "asis"}
panel_title <- function(label) {
  list(list(text = paste0("<b>", label, "</b>"), x = 0, y = 1.02, xref = "paper", yref = "paper",
            xanchor = "left", yanchor = "bottom", showarrow = FALSE, font = list(size = 12)))
}
link_axes <- function(sp, n) {
  ax <- paste0("xaxis", c("", 2:n))
  sp <- do.call(layout, c(list(sp), setNames(rep(list(list(matches = "x")), n - 1), ax[-1])))
  sp |> layout(margin = list(t = 30)) |> config(displaylogo = FALSE)
}

# ---- cyanotoxins, all sites ----
tox_labels <- c(mc = "Microcystin (ug/L)", stx = "Saxitoxin (ug/L)", atx = "Anatoxin-a (ug/L)", cyl = "Cylindrospermopsin (ug/L)")
site_order <- nc |> group_by(site) |> summarise(first = min(date)) |> arrange(first) |> pull(site)
site_cols  <- setNames(adjustcolor(unname(palette.colors(length(site_order), "Polychrome 36")), red.f = 0.75, green.f = 0.75, blue.f = 0.75), site_order)   # darkened so pale colors stay visible

tox_widget <- function() {
  legend_done <- character()   # sites that already have a legend entry (one per site, in the first panel where it has data)
  panels <- imap(tox_labels, function(label, par) {
    p <- plot_ly()
    for (s in site_order) {
      d <- nc |> filter(site == s) |> transmute(date, value = .data[[par]]) |> drop_na(value) |>
        mutate(txt = paste0("<b>", s, "</b><br>", format(date, "%d %b %Y"), "<br>", if_else(value == 0, "not detected", paste(signif(value, 3), "ug/L"))))
      for (det in c(TRUE, FALSE)) {   # detected (filled) first, so legend markers are filled whenever possible
        dd <- if (det) filter(d, value > 0) else filter(d, value == 0)
        if (nrow(dd) == 0) next
        show <- !(s %in% legend_done)
        if (show) legend_done <<- c(legend_done, s)
        p <- add_trace(p, data = dd, x = ~date, y = ~value, type = "scatter", mode = "markers", name = s, legendgroup = s,
                       showlegend = show, text = ~txt, hoverinfo = "text",
                       marker = list(color = site_cols[[s]], size = 8, symbol = if (det) "circle" else "circle-open",
                                     line = list(color = site_cols[[s]], width = 1.5)))
      }
    }
    p |> layout(annotations = panel_title(label), xaxis = list(title = "", tickfont = list(size = 9), tickformat = "%b %Y"),
                yaxis = list(title = "", tickfont = list(size = 9), rangemode = "tozero"))
  })
  sp <- subplot(unname(panels), nrows = 4, margin = c(0.04, 0.04, 0.05, 0.05), shareX = FALSE, titleY = FALSE)
  sp <- link_axes(sp, 4) |> layout(legend = list(title = list(text = "<b>Site</b>")))
  sp$height <- 850
  sp
}

# ---- nutrients, TLNWR1a_DH only ----
nut_labels <- c(tp = "Total P (ug/L)", tdp = "Total dissolved P (ug/L)", po4 = "Orthophosphate (ug/L)",
                tn = "Total N (ug/L)", nh3 = "Ammonia (ug/L)", n_ox = "Nitrate/nitrite (ug/L)")

nut_widget <- function() {
  panels <- imap(nut_labels, function(label, par) {
    d <- nc |> filter(site == "TLNWR1a_DH") |> transmute(date, value = .data[[par]]) |> drop_na(value) |> arrange(date) |>
      mutate(txt = paste0(format(date, "%d %b %Y"), "<br>", if_else(value == 0, "below detection (0)", paste(signif(value, 3), "ug/L"))))
    plot_ly(showlegend = FALSE) |>
      add_trace(data = d, x = ~date, y = ~value, type = "scatter", mode = "lines+markers", text = ~txt, hoverinfo = "text",
                line = list(color = "grey70", width = 1), marker = list(color = "steelblue", size = 8)) |>
      add_trace(data = filter(d, value == 0), x = ~date, y = ~value, type = "scatter", mode = "markers", text = ~txt, hoverinfo = "text",
                marker = list(color = "white", size = 9, line = list(color = "firebrick", width = 1.5))) |>
      layout(annotations = panel_title(label), xaxis = list(title = "", tickfont = list(size = 9), tickformat = "%b %Y"),
             yaxis = list(title = "", tickfont = list(size = 9), rangemode = "tozero"))
  })
  sp <- subplot(unname(panels), nrows = 3, margin = c(0.05, 0.05, 0.08, 0.06), shareX = FALSE, titleY = FALSE)
  sp <- link_axes(sp, 6)
  sp$height <- 700
  sp
}

tab_defs <- list(
  list(title = "Cyanotoxins (all sites)", note = "Microcystin, saxitoxin, anatoxin-a and cylindrospermopsin in each sample.", widget = tox_widget),
  list(title = "Nutrients (TLNWR1a_DH)", note = paste0("Lab nutrient results, the only site with nutrient data (", n_nut, " samples)."), widget = nut_widget)
)
for (tb in tab_defs) {
  writeLines(c("", "", paste0("### ", tb$title), "", tb$note, ""))
  print(htmltools::tagList(tb$widget()))
}
```

# 6. Flow and water level (context series)

```{r ancillary}
ctx <- bind_rows(
  read_txt("AR_Q") |> transmute(series = "Anderson Rose flow (Q; units not stated)", date = excel_date(date), value = as.numeric(q)),
  read_txt("Ady_Q") |> transmute(series = "Ady Canal flow (Q; units not stated)", date = excel_date(date), value = as.numeric(q)),
  read_txt("Unit 2 WSE") |> transmute(series = "Unit 2 water surface elevation (ft)", date = excel_date(date), value = as.numeric(wse_ft))
) |> drop_na(value)

ctx |>
  group_by(Series = series) |>
  summarise(First = min(date), Last = max(date), `Days with data` = n_distinct(date), .groups = "drop") |>
  kable()
```

These are context series for interpreting the sonde and cyano data (not water quality). Per the workbook, Ady flow comes from a USGS gage
(site 11509250), Anderson Rose flow from TID daily reports, and Unit 2 elevation from refuge staff. Flow units are not stated.
Unit 2 elevation is sparse before 2024 (`r sum(ctx$series == "Unit 2 water surface elevation (ft)" & year(ctx$date) < 2023)` reading before 2023, `r sum(ctx$series == "Unit 2 water surface elevation (ft)" & year(ctx$date) == 2023)` in 2023); the flow series only start in 2023 (Ady) and 2024 (Anderson Rose).
Hover for values, drag to zoom (all panels zoom together).

```{r ctx-plot}
ctx_panels <- imap(split(ctx, factor(ctx$series, levels = unique(ctx$series))), function(d, label) {
  d <- d |> arrange(date) |> mutate(txt = paste0(format(date, "%d %b %Y"), "<br><b>", signif(value, 4), "</b>"))
  plot_ly(showlegend = FALSE) |>
    add_trace(data = d, x = ~date, y = ~value, type = "scatter", mode = "lines+markers", text = ~txt, hoverinfo = "text",
              line = list(color = "steelblue", width = 1.2), marker = list(color = "steelblue", size = 3)) |>
    layout(annotations = panel_title(label), xaxis = list(title = "", tickfont = list(size = 9), tickformat = "%b %Y"),
           yaxis = list(title = "", tickfont = list(size = 9)))
})
w <- subplot(unname(ctx_panels), nrows = 3, margin = c(0.04, 0.04, 0.07, 0.05), shareX = FALSE, titleY = FALSE)
w <- link_axes(w, 3)
w$height <- 650
w
```

# 7. Things to note when using the data

- **Dissolved oxygen quality:** `r neg_do`% of continuous DO readings are negative (as low as -0.49 mg/L), almost all (99%) at `LKNWR_AI` and `TLNWR1a_LRI`, mostly July to October, and about 30% of readings at those two sites are below 1 mg/L. This looks like real low-oxygen conditions combined with a sensor offset near zero. DO saturation also reaches `r sprintf("%.0f", max(cont$do_sat, na.rm = TRUE))`% (`TLNWR1a_DH`), likely bloom-driven or a sensor artifact.
- **Sensor notes:** the Notes column records sonde swaps ("New sonde"), "Reference check" readings and algae fouling with wiper changes, which explain some jumps.
- **Site coverage differs:** the continuous sonde sites do not overlap in time, so sites and years cannot be compared season-for-season.
- **Non-detects:** `0` means not detected; detection limits are not provided.
- **Naming and locations:** site ids mix `-` and `_` (standardised to `_` here); `LKNWR_AI` and `LKNWR_SLBR` share the same coordinates; `LKNWR_NE` has no coordinates or description.
- **Discrete sonde readings:** the YSI tab also has about 30 one-off discrete readings (including top and bottom of the water column), which are not plotted.
