---
title: "Refuges Water Quality Data Exploration (Lower Klamath and Tule Lake NWR)"
author: "Badhia Yunes Katz"
date: "September 2026"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 9, fig.height = 4.5)
library(tidyverse)
library(readxl)
library(janitor)
library(lubridate)
library(leaflet)
library(knitr)
```

This report explores the "Refuges water quality data for Recovery" workbook
(`data-raw/skinner-wq-data/Refuges water quality data for Recovery.xlsx`) to understand its temporal coverage, data gaps and
collection points, and how it could be formatted like the other water quality objects in klamathWaterData.

The workbook covers Lower Klamath National Wildlife Refuge (LKNWR) and Tule Lake National Wildlife Refuge (TLNWR).
The tabs used in this report:

| Tab | Content |
|---|---|
| 2024+ sampling sites, 2021-2023 cyano sampling sites | Site descriptions and coordinates |
| Metadata | Column definitions |
| YSI | Sonde data: hourly continuous records plus a few discrete readings |
| Nutrient and cyano data | Discrete nutrient and cyanotoxin samples (2021-2026) |
| AR_Q, Ady_Q, Unit 2 WSE | Daily flow (Anderson Rose, Ady Canal) and Unit 2 water surface elevation |
| Notes on observed taxa, State of CA tabs | Reference material (not analysed here) |

```{r read}
f <- "data-raw/skinner-wq-data/Refuges water quality data for Recovery.xlsx"
excel_date <- function(x) as.Date(as.numeric(x), origin = "1899-12-30")
read_txt <- function(sheet) {
  read_excel(f, sheet = sheet, col_types = "text", .name_repair = "minimal") |>
    clean_names() |>
    mutate(across(everything(), ~ na_if(.x, "NA")))
}
fix_site <- function(x) str_replace(x, "-", "_")   # site ids mix "-" and "_" (e.g. TLNWR1a-NL vs TLNWR1a_NL)

sonde_cols <- c("temp", "p_h", "do_conc", "do_sat", "sp_cond", "tds", "sal", "depth", "bat", "cab")
nut_cols   <- c("tdp", "tp", "po4", "nh3", "n_ox", "tn", "pp", "per_pp", "n_p")
tox_cols   <- c("atx", "cyl", "mc", "stx")

ysi <- read_txt("YSI") |>
  mutate(date = excel_date(date), site = fix_site(site),
         across(all_of(sonde_cols), as.numeric))

nc <- read_txt("Nutrient and cyano data ") |>
  mutate(date = excel_date(date), site = fix_site(site),
         across(all_of(c(nut_cols, tox_cols)), as.numeric))

sites_2024 <- read_excel(f, sheet = "2024+ sampling sites", .name_repair = "minimal") |>
  clean_names() |> transmute(site = fix_site(site_name), description = site_description, coordinates)
sites_2021 <- read_excel(f, sheet = "2021-2023 cyano sampling sites", .name_repair = "minimal") |>
  clean_names() |> filter(!is.na(coordinates)) |>
  transmute(site = fix_site(site_id), description = NA_character_, coordinates)
sites_data <- nc |> filter(!is.na(coordinates)) |>
  distinct(site, .keep_all = TRUE) |> select(site, description = site_description, coordinates)
```

# 1. Collection points

Sites come from three places (the two site tabs and the coordinates recorded in the data), so they are combined here.
Site ids use both `-` and `_` (e.g. `TLNWR1a-NL` / `TLNWR1a_NL`); they were standardised to `_`.

```{r sites}
site_coords <- bind_rows(sites_2024, sites_2021, sites_data) |>
  distinct(site, .keep_all = TRUE) |>
  separate(coordinates, c("latitude", "longitude"), sep = ",", convert = TRUE) |>
  mutate(latitude = as.numeric(latitude), longitude = as.numeric(longitude))

all_dates <- bind_rows(ysi |> transmute(site, date, dataset = paste0("sonde (", str_to_lower(type), ")")),
                       nc  |> transmute(site, date, dataset = "nutrient/cyano"))

site_tbl <- all_dates |>
  group_by(site) |>
  summarise(sonde_continuous_days = n_distinct(date[dataset == "sonde (continuous)"]),
            sonde_discrete = sum(dataset == "sonde (discrete)"),
            nutrient_cyano_samples = sum(dataset == "nutrient/cyano"),
            first = min(date), last = max(date), .groups = "drop") |>
  left_join(site_coords, by = "site") |>
  mutate(refuge = if_else(str_starts(site, "LKNWR"), "Lower Klamath", "Tule Lake")) |>
  arrange(refuge, site) |>
  select(refuge, site, description, latitude, longitude, everything())

kable(site_tbl)
```

- **`r nrow(site_tbl)` sites** in total: `r sum(site_tbl$refuge == "Lower Klamath")` at Lower Klamath and `r sum(site_tbl$refuge == "Tule Lake")` at Tule Lake.
- **Continuous sonde** sites are `LKNWR_NE` (2023), `LKNWR_AI` (2024-2025), `TLNWR1a_DH` (2024) and `TLNWR1a_LRI` (2024-2025).
- **Cyano/nutrient sampling** was routine at Sheepy Lake (`LKNWR_SLBR`, `LKNWR_WSL`) and Tule Lake sump 1A/1B (`_BR`, `_NL`, `_SW`, `_NW`) in 2021-2023. After April 2023 only a few sites were sampled, mostly during blooms (`TLNWR1a_DH` continues to 2026).
- Sites such as `TLNWR1a_EN`, `_SPR`, `_STN` and the `bot` (bottom) samples are one-off readings; `LKNWR_NE` has no coordinates or description.
- `LKNWR_AI` and `LKNWR_SLBR` share the same coordinates in the site tab.

```{r map}
site_tbl |>
  drop_na(latitude, longitude) |>
  mutate(has_sonde = sonde_continuous_days > 0) |>
  leaflet() |>
  addProviderTiles("Esri.WorldImagery") |>
  addCircleMarkers(~longitude, ~latitude, radius = 7, color = "white", weight = 1,
                   fillColor = ~if_else(has_sonde, "orange", "steelblue"), fillOpacity = 0.9,
                   label = ~site,
                   popup = ~paste0("<b>", site, "</b><br>", coalesce(description, ""),
                                   "<br>Continuous sonde days: ", sonde_continuous_days,
                                   "<br>Nutrient/cyano samples: ", nutrient_cyano_samples)) |>
  addLegend("bottomright", colors = c("orange", "steelblue"),
            labels = c("Continuous sonde site", "Discrete sampling only"), opacity = 1)
```

Sites without coordinates: `r paste(site_tbl$site[is.na(site_tbl$latitude)], collapse = ", ")`.

# 2. Sonde data (YSI tab)

## 2.1 Temporal coverage

```{r deployments}
cont <- ysi |> filter(type == "Continuous")

deployments <- cont |>
  group_by(site, year) |>
  summarise(first = min(date), last = max(date),
            readings = n(), days_with_data = n_distinct(date),
            span_days = as.numeric(max(date) - min(date)) + 1, .groups = "drop") |>
  mutate(missing_days_in_span = span_days - days_with_data)

kable(deployments, caption = "Continuous sonde deployments")
```

```{r sonde-timeline}
ysi |>
  distinct(site, type, date) |>
  mutate(site = fct_reorder(site, date, .fun = min)) |>
  ggplot(aes(date, site, colour = type)) +
  geom_point(shape = "|", size = 4) +
  scale_colour_brewer(palette = "Dark2") +
  guides(colour = guide_legend(override.aes = list(size = 4))) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  labs(x = NULL, y = NULL, colour = "Record type", title = "Days with sonde data") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "top")
```

- Continuous records run from **`r min(cont$date)`** to **`r max(cont$date)`** and are hourly.
- Records are seasonal, and each site has its own period: `LKNWR_NE` May-Sep 2023, `TLNWR1a_DH` May-Aug 2024 (sonde stopped 16 Aug 2024), `TLNWR1a_LRI` Aug 2024-Oct 2025 and `LKNWR_AI` May 2024-Oct 2025.
- Winter gap: `LKNWR_AI` has no data from 12 Nov 2024 to 27 May 2025 (196 days), and `TLNWR1a_LRI` has no data from 26 Nov 2024 to 27 May 2025.
- After the `TLNWR1a_DH` sonde ended in Aug 2024, that site only has discrete readings (4-6 per year).
- There are no continuous data for any site in 2021, 2022 or 2026.

## 2.2 Parameter completeness and value ranges

```{r completeness}
cont |>
  mutate(deployment = paste(site, year, sep = " / ")) |>
  select(deployment, all_of(sonde_cols)) |>
  pivot_longer(-deployment, names_to = "parameter", values_to = "value") |>
  group_by(deployment, parameter) |>
  summarise(pct = 100 * mean(!is.na(value)), .groups = "drop") |>
  ggplot(aes(parameter, deployment, fill = pct)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = round(pct)), size = 3) +
  scale_fill_viridis_c(limits = c(0, 100), name = "% present") +
  labs(x = NULL, y = NULL, title = "% of continuous readings with a value") +
  theme_minimal()
```

```{r ranges}
cont |>
  select(all_of(c("temp", "p_h", "do_conc", "do_sat", "sp_cond", "depth"))) |>
  pivot_longer(everything(), names_to = "parameter", values_to = "value") |>
  group_by(parameter) |>
  summarise(n = sum(!is.na(value)), min = min(value, na.rm = TRUE), median = median(value, na.rm = TRUE),
            max = max(value, na.rm = TRUE), n_negative = sum(value < 0, na.rm = TRUE)) |>
  kable(digits = 2)
```

- Temperature, pH, DO and conductivity are complete in every deployment. The battery (`bat`, `cab`) columns are empty for `LKNWR_NE`.
- **Dissolved oxygen has a data-quality issue**: 13% of continuous DO readings are negative (down to -0.49 mg/L), all at `LKNWR_AI` and `TLNWR1a_LRI` in the summer months, and about a third of readings at those sites are below 1 mg/L. This looks like real low-oxygen conditions combined with a sensor offset near zero. DO saturation also reaches 341% (`TLNWR1a_DH`), likely bloom-driven or a sensor artifact.
- The Notes column records sonde swaps ("New sonde"), "Reference check" readings, and algae fouling with wiper changes, which explain some jumps.

```{r ts, fig.height = 7}
cont |>
  mutate(doy_date = as.Date(paste0("2000-", format(date, "%m-%d"))), year = factor(year(date))) |>
  select(doy_date, year, site, temp, do_conc, p_h) |>
  pivot_longer(c(temp, do_conc, p_h), names_to = "parameter", values_to = "value") |>
  drop_na(value) |>
  ggplot(aes(doy_date, value, colour = site)) +
  geom_point(size = 0.3, alpha = 0.4) +
  facet_wrap(~parameter, scales = "free_y", ncol = 1) +
  scale_colour_brewer(palette = "Dark2") +
  guides(colour = guide_legend(override.aes = list(size = 4, alpha = 1))) +
  scale_x_date(date_labels = "%b") +
  labs(x = NULL, y = NULL, colour = "Site", title = "Hourly sonde readings by day of year") +
  theme_minimal() +
  theme(legend.position = "top", legend.text = element_text(size = 11))
```

TODO this plot could be improved to make the data easier to see. Options: facet by site, plot daily mean with min/max ribbons instead of hourly points, or use a larger point size with less overplotting.

# 3. Nutrient and cyanotoxin samples

## 3.1 Temporal coverage

```{r nc-timeline}
nc |>
  mutate(refuge = if_else(str_starts(site, "LKNWR"), "Lower Klamath", "Tule Lake"),
         site = fct_reorder(site, date, .fun = min)) |>
  ggplot(aes(date, site, colour = refuge)) +
  geom_point(alpha = 0.7, size = 2) +
  scale_colour_brewer(palette = "Dark2") +
  guides(colour = guide_legend(override.aes = list(size = 4, alpha = 1))) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  labs(x = NULL, y = NULL, colour = "Refuge", title = "Nutrient / cyano sampling dates") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "top")
```

```{r nc-year}
nc |> count(year = year(date), name = "samples") |> kable(caption = "Samples per year")
```

- **`r nrow(nc)` samples** from **`r min(nc$date)`** to **`r max(nc$date)`**.
- Sampling was dense in 2021-2022 (93 and 114 samples) and ended for most routine sites in April 2023 (24 samples in 2023). From 2024 there are only 5-10 samples a year, mostly bloom-driven.
- Notes explain some skipped sites, e.g. sites not sampled because two consecutive toxin results were non-detect.

## 3.2 Which parameters were measured

```{r nc-complete}
nc |>
  select(date, site, all_of(c(nut_cols, tox_cols))) |>
  pivot_longer(-c(date, site), names_to = "parameter", values_to = "value") |>
  group_by(parameter) |>
  summarise(n_values = sum(!is.na(value)),
            n_sites = n_distinct(site[!is.na(value)]),
            first = min(date[!is.na(value)]), last = max(date[!is.na(value)]),
            n_zero = sum(value == 0, na.rm = TRUE)) |>
  kable(caption = "Non-missing values by parameter (n_zero = non-detect)")
```

- **Cyanotoxins (ATX, CYL, MC, STX)** were measured at most sampling events, with `0` meaning non-detect.
- **Nutrients (TDP, TP, PO4, NH3, NOx, TN)** were only measured at `TLNWR1a_DH`, and only about a dozen times (2024-2026). The nutrient columns are empty for the 2021-2023 routine cyano sites.
- As in the other workbooks, a nutrient value of `0` means below the detection limit, not a true zero.

# 4. Flow and water level tabs

```{r ancillary}
map_dfr(c(AR_Q = "AR_Q", Ady_Q = "Ady_Q", `Unit 2 WSE` = "Unit 2 WSE"), function(sh) {
  d <- read_txt(sh) |> mutate(date = excel_date(date))
  tibble(days = nrow(d), first = min(d$date), last = max(d$date),
         missing_days = as.numeric(max(d$date) - min(d$date)) + 1 - n_distinct(d$date))
}, .id = "tab") |>
  mutate(content = c("Anderson Rose flow (Q), daily", "Ady Canal flow (Q), daily", "Unit 2 water surface elevation (ft)")) |>
  kable()
```

- These are context series for interpreting the sonde and cyano data, not water quality.
- Per the Metadata tab, Ady flow is from a USGS gage (site 11509250), Anderson Rose flow comes from TID daily reports, and Unit 2 water surface elevation was provided by refuge staff.
- Units are not stated for the flow columns. The USGS Ady record could instead be pulled directly if needed for `flow_data`.

# 5. Formatting for klamathWaterData

Existing continuous-data objects (`temperature_data`, `do_data`, `ph_data`) hold **daily** `mean` / `minimum` / `maximum`
values with columns `location, gage_name, gage_id, variable_name, value, unit, statistic, date`, plus a `*_gage` table.
Draft conversion of the **continuous** sonde data (days with at least 20 hourly readings; negative DO dropped):

```{r package-format}
param_map <- tribble(
  ~col,      ~variable_name,      ~unit,
  "temp",    "temperature",       "celsius",
  "do_conc", "dissolved oxygen",  "mg/L",
  "p_h",     "pH",                "std unit"
)

site_info <- site_tbl |>
  mutate(location  = if_else(refuge == "Lower Klamath", "lower klamath national wildlife refuge",
                             "tule lake national wildlife refuge"),
         gage_name = str_to_lower(coalesce(description, site)),
         gage_id   = paste0("REFUGES-", site)) |>
  select(site, location, gage_name, gage_id, latitude, longitude)

refuges_wq_data <- cont |>
  select(site, date, all_of(param_map$col)) |>
  pivot_longer(all_of(param_map$col), names_to = "col", values_to = "value") |>
  filter(!is.na(value), !(col == "do_conc" & value < 0)) |>
  group_by(site, date, col) |>
  summarise(n = n(), mean = mean(value), minimum = min(value), maximum = max(value), .groups = "drop") |>
  filter(n >= 20) |>
  pivot_longer(c(mean, minimum, maximum), names_to = "statistic", values_to = "value") |>
  left_join(param_map, by = "col") |>
  left_join(site_info, by = "site") |>
  select(location, gage_name, gage_id, variable_name, value, unit, statistic, date) |>
  as.data.frame()

refuges_wq_gage <- site_info |>
  filter(site %in% cont$site) |>
  transmute(gage_name, gage_id, agency = NA_character_, latitude, longitude,
            river_mile = NA_real_, huc8 = NA_character_, location)

refuges_wq_data |>
  filter(statistic == "mean") |>
  count(gage_id, variable_name, name = "n_days") |>
  pivot_wider(names_from = variable_name, values_from = n_days) |>
  kable(caption = "Daily records per site and variable")
```

`LKNWR_NE` has no coordinates, so it cannot get a gage location without the provider.

# 6. Summary and open questions

**Coverage**

- Four sites have continuous hourly sonde data, all seasonal and non-overlapping in period: `LKNWR_NE` (2023), `LKNWR_AI` (2024-2025), `TLNWR1a_DH` (May-Aug 2024) and `TLNWR1a_LRI` (Aug 2024-Oct 2025). There is a winter gap (Nov-May) at both 2024-2025 sites.
- Nutrient/cyanotoxin samples span 2021-2026 at 14 sites; they are dense in 2021-2023, then sparse. Nutrients are only available at `TLNWR1a_DH`.
- Some sonde values are implausible (13% of DO readings negative at two sites, DO saturation up to 341%) and need a QC pass.

**Decisions needed before adding to the package**

1. **Sonde summarisation** - daily mean/min/max (as drafted, with a 20-reading minimum) or keep hourly values?
2. **Which variables** - temperature, DO and pH map to existing objects; conductivity, TDS, salinity and depth would need new objects.
3. **Nutrients and cyanotoxins** - the package has no nutrient or HAB objects yet. Add new objects, or leave out? Nutrients exist for only one site.
4. **Discrete sonde readings** - include the 30 or so discrete YSI rows (e.g. top/bottom of water column, `Eagle's Nest`) or continuous data only?
5. **QC** - how to treat negative DO (dropped in the draft above, which removes about 170 site-days (of ~750) at `LKNWR_AI` and `TLNWR1a_LRI`; alternatively set to 0 or flag) and other implausible values; confirm whether the provider has already QC'd the data ("Reference check" and "New sonde" rows are flagged in Notes).
6. **Non-detects** - keep `0` as non-detect (with a flag) or set to missing? Detection limits are not provided.
7. **`location` / `agency` / `huc8`** - the refuge names are placeholders; confirm data owner (USFWS refuge staff / partner lab) and HUC8.
8. **Missing site information** - `LKNWR_NE` has no coordinates or description; site ids mix `-` and `_`; `LKNWR_AI` and `LKNWR_SLBR` share coordinates.
9. **Flow and water level tabs** - Ady Q, Anderson Rose Q and Unit 2 WSE could feed `flow_data` / `water_level_data`; confirm units and whether to include them.
