---
title: "HAB data Exploration "
author: "Badhia Yunes Katz"
date: "August 2025"
# output: rmarkdown::github_document
---


```{r include=FALSE}
library(tidyverse)
library(dataRetrieval)
library(leaflet)
library(janitor)
library(dplyr)
library(ggplot2)
library(DT)         
library(tidyr)
library(scales)
library(leaflet)
```


This report is a **one-stop overview** of harmful algal bloom (HAB) data and practice in the **Klamath Basin** (southern Oregon & northern California). It covers:

- **What HAB parameters matter** and the **thresholds** used in health guidance (with links to sources)
- **What data exists** in the **Water Quality Portal (WQX)** for the basin
- A **reproducible R workflow** to pull those data for the last ~10 years, flag likely HAB events, and map them
- **Caveats** and tips for interpreting results

The goal: let any reader understand the **what/why** and **see the data**.

---

# 2. Basin context (very brief)

The Klamath system includes **Upper Klamath Lake**, the **Klamath River**, and major tributaries. HABs occur as:

- **Planktonic blooms** (e.g., *Microcystis*) → often produce **microcystins** (toxin)
- **Benthic cyanobacterial mats** (e.g., *Microcoleus*) → can produce **anatoxin-a**, especially in river reaches

**Context resources**  
- Oregon Health Authority (OHA) HAB program (advisories, criteria, program logic):  
  https://www.oregon.gov/oha/PH/HEALTHYENVIRONMENTS/RECREATION/HARMFULALGAEBLOOMS/Pages/
- USGS Klamath studies overview:  
  https://www.usgs.gov/centers/oregon-water-science-center/science/upper-klamath-basin-studies-and-data-collection

---

# 3. HAB parameters and **why** they matter (with thresholds & sources)

These are the most decision-relevant parameters for **recreational waters**. We use them to classify **toxic events** and **bloom evidence**.

| Parameter | Threshold (recreation context) | Why it matters | Sources (working links) |
|---|---:|---|---|
| **Microcystin** | **≥ 8 µg/L** | Most common cyanotoxin; core basis for advisories | EPA 2019 Criteria (PDF): https://www.epa.gov/sites/default/files/2019-05/documents/hh-rec-criteria-habs-document-2019.pdf ; OHA 2024 Guidance (PDF): https://www.oregon.gov/oha/PH/HEALTHYENVIRONMENTS/RECREATION/HARMFULALGAEBLOOMS/Documents/Advisory%20Guidelines%20for%20Harmful%20Cyanobacteria%20Blooms%20in%20Recreational%20Waters.2024_Final.pdf |
| **Cylindrospermopsin** | **≥ 15 µg/L** | Cyanotoxin; used in EPA/OHA frameworks | EPA 2019 Criteria (PDF): https://www.epa.gov/sites/default/files/2019-05/documents/hh-rec-criteria-habs-document-2019.pdf |
| **Anatoxin-a** | **≥ 15 µg/L** | Neurotoxin; common in **benthic mats** | OHA 2024 Guidance (PDF): https://www.oregon.gov/oha/PH/HEALTHYENVIRONMENTS/RECREATION/HARMFULALGAEBLOOMS/Documents/Advisory%20Guidelines%20for%20Harmful%20Cyanobacteria%20Blooms%20in%20Recreational%20Waters.2024_Final.pdf |
| **Saxitoxin** | **≥ 8 µg/L** | Neurotoxin; advisory value | OHA 2024 Guidance (PDF): link above |
| **Cyanobacteria (cells)** | **≥ 100,000 cells/mL** | WHO alert-level trigger; risk management | WHO 2nd ed. *Toxic Cyanobacteria in Water* (landing): https://www.who.int/publications/m/item/toxic-cyanobacteria-in-water---second-edition ; Full open PDF: https://library.oapen.org/bitstream/20.500.12657/47047/1/9781000262025.pdf |
| **Chlorophyll a** | **≈ 20–50 µg/L** when cyano-dominated; **use ≥ 30 µg/L** for conservative screening | Biomass proxy; supports bloom identification | WHO 2nd ed., Ch. 5 (open PDF): link above |
| **Phycocyanin** | **Site-specific (no universal cut-point)**; use **local 90th percentile** as screening | Cyanobacteria pigment (sensor reading); good early warning when calibrated to cells/toxins | EPA 2019 Monitoring Guidance (PDF): https://www.epa.gov/sites/default/files/2019-09/documents/recommend-cyano-rec-water-2019-update.pdf ; Review: Almuhtaram et al., 2021 (overview of sensors and variability) https://www.sciencedirect.com/science/article/pii/S1470160X21011079 |

> **Benthic vs planktonic:** In the Klamath River, **benthic mats** can drive **anatoxin-a** even when the water column seems clear—plan monitoring accordingly (see e.g., Genzoli et al., 2024, *Freshwater Science*).

---

# 4. What data is available (and where)

- **Water Quality Portal (WQX/NWIS/STORET):** In-situ measurements reported by agencies/tribes/utilities. We will pull the 7 parameters above for the Klamath Basin. https://www.waterqualitydata.us/
- **OHA HAB advisories (issue/lift dates):** Official public-health decisions for Oregon waters. https://www.oregon.gov/oha/PH/HEALTHYENVIRONMENTS/RECREATION/HARMFULALGAEBLOOMS/Pages/Blue-GreenAlgaeAdvisories.aspx
- **KBMP (Klamath Basin Monitoring Program) Blue-Green Algae Tracker:** Reach-level flags where thresholds are exceeded; useful spatial context. https://kbmp.net/maps-and-data
- **CyAN (EPA/NOAA/USGS) satellite index:** Weekly cyanobacteria index for large waters (e.g., Upper Klamath Lake). Public viewer: https://experience.arcgis.com/experience/01f798892ecb48d8ac99ef9b4ebe37c8

This report **focuses on WQX** and provides a defensible way to call events.

---

# 5. How we’ll classify “events” in this report

- **Toxic HAB event (primary evidence):** any sample with **Microcystin ≥ 8 µg/L** (or other cyanotoxin ≥ its threshold).
- **Bloom evidence (supporting):** **Cyanobacteria ≥ 100,000 cells/mL**, **Chlorophyll-a ≥ 30 µg/L** (screening), or **Phycocyanin ≥ 90th percentile** locally.
- **Interpretation tip:** If toxins are absent but **benthic mats** are present, consider **anatoxin-a** detection and site context.

---

# 6. Reproducible analysis (pulling WQX, flagging, mapping)

> This code pulls the last ~10 years for **Klamath Basin (HUC2 = 180102)**. You can change the dates if needed.



The workflow:

1. Query WQP for relevant parameters  
2. Summarize maximum values per year and site  
3. Identify likely bloom events  
4. Map bloom locations and provide a summary table  


```{r }
# Define Klamath Basin HUC8 codes
huc_code <- "180102"

hab_characteristics <- c("Phycocyanin", "Chlorophyll a","Microcystin","Cyanobacteria", "Anatoxin-A", "Saxitoxin", "Cylindrospermopsin")

# Set date range
start_date <- "2014-01-01"
end_date <- "2025-12-31"
```

```{r}
# Pull result data
wqx_results <- map(hab_characteristics, function(cn) {
  message("Querying WQP for: ", cn)
  tryCatch(
    readWQPdata(
      huc = huc_code,
      characteristicName = cn,
      startDate = start_date,
      endDate   = end_date,
      sampleMedia = "Water"
    ),
    error = function(e) { warning("WQP error for ", cn, ": ", e$message); NULL }
  )
}) |> compact()

stopifnot(length(wqx_results) > 0)    # basic guard

harmonize_wqx <- function(df) {
  df <- janitor::clean_names(df)

  # Ensure key columns exist
  needed <- c(
    "result_measure_value",
    "result_measure_measure_unit_code",
    "activity_start_date",
    "monitoring_location_identifier",
    "monitoring_location_name",
    "characteristic_name",
    "latitude_measure",
    "longitude_measure"
  )
  for (nm in needed) if (!nm %in% names(df)) df[[nm]] <- NA_character_

  # Force consistent types BEFORE binding; parse later
  df |>
    mutate(
      result_measure_value = as.character(result_measure_value),
      result_measure_measure_unit_code = as.character(result_measure_measure_unit_code),
      activity_start_date = as.character(activity_start_date),
      latitude_measure = as.character(latitude_measure),
      longitude_measure = as.character(longitude_measure),
      characteristic_name = as.character(characteristic_name),
      monitoring_location_identifier = as.character(monitoring_location_identifier),
      monitoring_location_name = as.character(monitoring_location_name)
    )
}

# Apply harmonization to each chunk, then bind
all_results <- wqx_results |>
  purrr::map(harmonize_wqx) |>
  dplyr::bind_rows()

# Now safely coerce to usable types
all_results <- all_results |>
  mutate(
    date      = as.Date(activity_start_date),
    value_num = suppressWarnings(readr::parse_number(result_measure_value)),
    latitude_measure  = suppressWarnings(as.numeric(latitude_measure)),
    longitude_measure = suppressWarnings(as.numeric(longitude_measure))
  )

# Pull site metadata to get names and coordinates
site_info <- whatWQPsites(huc = huc_code) |> clean_names()

# Join & prepare fields
dat <- all_results |>
  left_join(site_info, by = "monitoring_location_identifier")

glimpse(dat)

```

```{r}
# Toxins (EPA 2019; OHA 2024)
thr_microcystin        <- 8    # µg/L
thr_cylindrospermopsin <- 15   # µg/L
thr_anatoxin_a         <- 15   # µg/L
thr_saxitoxin          <- 8    # µg/L

# Supporting indicators (WHO/literature)
thr_cyano_cells <- 100000      # cells/mL
thr_chla        <- 30          # µg/L (conservative screening)

# Phycocyanin "elevated": local 90th percentile (screening)
pc90 <- dat |>
  filter(characteristic_name == "Phycocyanin") |>
  summarise(pc90 = quantile(value_num, 0.90, na.rm = TRUE)) |>
  pull(pc90)
if (length(pc90) == 0 || is.na(pc90)) pc90 <- Inf

unit_safe <- function(x) ifelse(is.na(x) | x=="", "", x)

# Compute a Phycocyanin 90th percentile (across available pc data)
pc90 <- dat |>
  dplyr::filter(characteristic_name == "Phycocyanin", is.finite(value_num)) |>
  dplyr::summarise(pc90 = stats::quantile(value_num, 0.90, na.rm = TRUE)) |>
  dplyr::pull(pc90)
if (length(pc90) == 0 || is.na(pc90)) pc90 <- Inf

# Helper predicates to guard by unit where appropriate
is_ugL <- function(u) grepl("^\\s*ug/?L\\s*$", u, ignore.case = TRUE)  # µg/L variants
is_cells_ml <- function(u) grepl("cells\\s*/\\s*mL", u, ignore.case = TRUE)


hab_events <- dat |>
  dplyr::mutate(
    # Toxins: only compare when reported in µg/L
    toxic_event = dplyr::case_when(
      characteristic_name == "Microcystin" &
        is_ugL(result_measure_measure_unit_code) &
        value_num >= thr_microcystin ~ TRUE,
      characteristic_name == "Cylindrospermopsin" &
        is_ugL(result_measure_measure_unit_code) &
        value_num >= thr_cylindrospermopsin ~ TRUE,
      characteristic_name == "Anatoxin-a" &
        is_ugL(result_measure_measure_unit_code) &
        value_num >= thr_anatoxin_a ~ TRUE,
      characteristic_name == "Saxitoxin" &
        is_ugL(result_measure_measure_unit_code) &
        value_num >= thr_saxitoxin ~ TRUE,
      TRUE ~ FALSE
    ),
    # Supporting indicators with unit guards
    bloom_support = dplyr::case_when(
      characteristic_name == "Cyanobacteria" &
        is_cells_ml(result_measure_measure_unit_code) &
        value_num >= thr_cyano_cells ~ TRUE,
      characteristic_name == "Chlorophyll a" &
        is_ugL(result_measure_measure_unit_code) &
        value_num >= thr_chla ~ TRUE,
      characteristic_name == "Phycocyanin" &
        is.finite(value_num) & value_num >= pc90 ~ TRUE,
      TRUE ~ FALSE
    ),
    event = toxic_event | bloom_support
  ) |>
  dplyr::filter(event) |> 
  dplyr::select(
    date, monitoring_location_identifier, 
    monitoring_location_name.y,
    characteristic_name, value_num, result_measure_measure_unit_code,
    latitude_measure.y, longitude_measure.y, toxic_event, bloom_support
  )

# Quick preview
head(hab_events, 10)
```

Yearly context: per-site maxima

```{r echo=FALSE}
summary_table <- dat |> 
  mutate(year = year(date)) |> 
  group_by(monitoring_location_identifier, monitoring_location_name.y,
           characteristic_name, year) |> 
  summarise(samples = dplyr::n(),
            max_value = suppressWarnings(max(value_num, na.rm = TRUE)),
            units = dplyr::first(result_measure_measure_unit_code),
            .groups   = "drop") |> 
  arrange(desc(max_value))

head(summary_table, 10)
```


```{r echo=FALSE}
ggplot(summary_table, aes(x = year, y = max_value)) +
  geom_point() +
  facet_wrap(~ characteristic_name, scales = "free_y") +
  labs(
    title = "Maximum HAB-Related Values by Year (per parameter)",
    x = "Year", y = "Max value (units vary by parameter)"
  )

```


```{r echo=FALSE}
pal <- colorFactor(
  palette = c("darkred","darkgreen","blue","orange","purple","brown","goldenrod"),
  domain  = hab_events$characteristic_name
)

# leaflet(hab_events) |> 
#   addTiles() |> 
#   addCircleMarkers(
#     lng   = ~longitude_measure.y + runif(n(), -0.0005, 0.0005),  # small jitter
#     lat   = ~latitude_measure.y  + runif(n(), -0.0005, 0.0005),
#     color = ~pal(characteristic_name),
#     label = ~HTML(sprintf(
#       "<b>%s</b><br><b>Date:</b> %s<br><b>%s:</b> %s %s",
#       htmlEscape(monitoring_location_name.y),
#       format(date, "%Y-%m-%d"),
#       htmlEscape(characteristic_name),
#       number(value_num, accuracy = 0.01),
#       htmlEscape(unit_safe(result_measure_measure_unit_code))
#     )),
#     radius = 6, fillOpacity = 0.8
#   ) |> 
#   addLegend("bottomright", pal = pal, values = ~characteristic_name, title = "HAB Parameter")

```


old code below

```{r eval=FALSE, include=FALSE}
all_results <- data.frame()

for (char_name in hab_characteristics) {
  message("Querying for: ", char_name)

  result <- tryCatch({
    readWQPdata(
      huc = huc_code,
      characteristicName = char_name,
      startDate = start_date,
      endDate = end_date,
      sampleMedia = "Water"
    )
  }, error = function(e) {
    warning(paste("Error retrieving", char_name, ":", e$message))
    return(NULL)
  })

  if (!is.null(result) && nrow(result) > 0) {
    all_results <- rbind(all_results, result)
  }
}

# cleaning column names
all_results <- all_results |>
  clean_names()
```

```{r eval=FALSE, include=FALSE}
site_info <- whatWQPsites(huc = huc_code) |> clean_names()

data <- all_results |>
  left_join(site_info, by = "monitoring_location_identifier") |>
  mutate(
    date = as.Date(activity_start_date),
    value_num = as.numeric(result_measure_value),
    latitude_measure  = as.numeric(latitude_measure),
    longitude_measure = as.numeric(longitude_measure))

summary_table <- data |>
  mutate(year = year(date)) |>
  group_by(monitoring_location_identifier, monitoring_location_name,
           characteristic_name, year) |>
  summarise(
    samples   = n(),
    max_value = max(value_num, na.rm = TRUE),
    units = result_measure_measure_unit_code,
    .groups = "drop") |>
  arrange(desc(max_value))

head(summary_table, 10)

```

```{r eval=FALSE, include=FALSE}
# Toxin thresholds (EPA 2019 / OHA 2024)
thr_microcystin      <- 8    # µg/L
thr_cylindrospermopsin <- 15 # µg/L
thr_anatoxin_a       <- 15   # µg/L (OHA RUV)
thr_saxitoxin        <- 8    # µg/L (OHA RUV)

# Supporting indicators (WHO / literature)
thr_cyano_cells <- 100000    # cells/mL
thr_chla <- 30        # µg/L (screening, conservative)

# Phycocyanin local "elevated" signal = 90th percentile (across dataset).
pc90 <- data |>
  filter(characteristic_name == "Phycocyanin") |>
  summarise(pc90 = quantile(value_num, 0.90, na.rm = TRUE)) |>
  pull(pc90)
if (length(pc90) == 0 || is.na(pc90)) pc90 <- Inf

# Helper: unit-safe string
unit_safe <- function(x) ifelse(is.na(x) | x=="", "", x)

```



### Summary of Annual Maximum Values

```{r include=FALSE}
# exploring high values
# summary_table <- all_results |> 
#   mutate(Year = year(ActivityStartDate)) |> 
#   group_by(MonitoringLocationIdentifier, CharacteristicName, Year) |> 
#   summarize(
#     Samples = n(),
#     MaxValue = max(as.numeric(ResultMeasureValue), na.rm = TRUE),
#     .groups = "drop") |> 
#   arrange(desc(MaxValue))
# 
# head(summary_table, 10)  # highest 10
```


### Likely Bloom Events

```{r echo=FALSE}
# likely bloom conditions based on research
# likely_blooms <- summary_table |> 
#   filter((CharacteristicName == "Microcystin" & MaxValue >= 8) |
#            (CharacteristicName == "Chlorophyll a" & MaxValue >= 30) |
#            (CharacteristicName == "Cyanobacteria" & MaxValue >= 100000))
# 
# ggplot(summary_table, aes(x = Year, y = MaxValue, color = CharacteristicName)) +
#   geom_point() +
#   facet_wrap(~ CharacteristicName, scales = "free_y") +
#   labs(title = "Maximum HAB-Related Values by Year",
#        y = "Max Value",
#        x = "Year")

```

### Likely HAB Events with Location

```{r eval=FALSE, include=FALSE}
# pulling site info 
site_info <- whatWQPsites(huc = "180102")

# join site info to data
all_results_with_sites <- left_join(all_results, site_info,
                                    by = "MonitoringLocationIdentifier")

all_results_with_sites <- all_results_with_sites |> 
  clean_names() |> 
  mutate(latitude_measure = as.numeric(latitude_measure),
         longitude_measure = as.numeric(longitude_measure)) |>
  glimpse()
  

hab_events <- all_results_with_sites |> 
  filter(characteristic_name %in% c("Microcystin", "Chlorophyll a", "Phycocyanin", "Cyanobacteria")) |> 
  mutate(value = as.numeric(result_measure_value),
         date = as.Date(activity_start_date),
         hab_likely = case_when(
           characteristic_name == "Microcystin" & value >= 8 ~ TRUE,
           characteristic_name == "Chlorophyll a" & value >= 30 ~ TRUE,
           characteristic_name == "Cyanobacteria" & value >= 100000 ~ TRUE,
           characteristic_name == "Phycocyanin" & value >= quantile(value, 0.90, na.rm = TRUE) ~ TRUE,
           TRUE ~ FALSE)) |> 
  filter(hab_likely == TRUE) |> 
  select(date, characteristic_name, value, result_depth_height_measure_measure_unit_code,
         monitoring_location_name, latitude_measure, longitude_measure)

habs_palette <- colorFactor(
  palette = c("darkred", "darkgreen", "blue", "orange"),
  domain = hab_events$characteristic_name)

hab_events_clean <- hab_events |> 
   mutate(latitude_measure = jitter(latitude_measure, factor = .07), #TODO figure out jitter
         longitude_measure = jitter(longitude_measure, factor = .07))

leaflet(hab_events_clean) |> 
  addTiles() |> 
  addCircleMarkers(lng = ~longitude_measure,
                   lat = ~latitude_measure,
                   color = ~habs_palette(characteristic_name),
                   label = ~paste0(monitoring_location_name,
                                  "<br><b>Date:</b>", date,
                                  "<br><b>", characteristic_name, ":</b> ", value,
                                  result_depth_height_measure_measure_unit_code),
                   radius = 6,
                   fillOpacity = 0.8) |> 
    addLegend("bottomright",
              pal = habs_palette,
              values = ~characteristic_name,
              title = "HAB Type")
```

### Summary Table of HAB Events

```{r eval=FALSE, include=FALSE}
# summary table
hab_summary_table <- hab_events |> 
  mutate(
    hab_type = characteristic_name,
    site = monitoring_location_name,
    value = round(value, 2),
    units = result_depth_height_measure_measure_unit_code,
    latitude = latitude_measure,
    longitude = longitude_measure) |> 
  select(date, site, hab_type, value) |> 
  arrange(desc(date))

hab_summary_table

```

